---
title: "Farm management diversification — exploratory summary"
subtitle: "Figures from `Scripts/02_exploratory.R`"
date: today
format:
  pdf:
    toc: false
    fig-pos: "H"
    geometry: margin=1in
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

## Regenerate every figure object (p_distributions, p_corr, ... p_std) and the
## Ecoregion_variance table from the analysis script itself, so this report can
## never drift from the code. Fast -- no iNEXT bootstrap here.
source("Scripts/02_exploratory.R")

library(knitr)

## Drop the on-plot title/subtitle for the report -- the section heading and the
## paragraph under each figure carry that text, and the long subtitles clip at
## the PDF page width
bare <- function(p) p + labs(title = NULL, subtitle = NULL)
```

This report summarises the exploratory plots of the four farm management
diversification indices (land use, water management, pasture management, and an
overall "all practices" index) on the `r nrow(Farm_div_matched)` SCR farms that
have an associated bird-diversity estimate. It has two parts: (1) the indices on
their own, and (2) how they — and bird diversity, and the main farm covariates —
vary among the five ecoregions.

# The indices on their own

## Distribution of each index

```{r}
#| label: fig-distributions
#| fig-width: 8
#| fig-height: 6

bare(p_distributions)
```

Each index runs 0–1. Land use and "all practices" are spread fairly evenly across
the range; water and pasture management pile up at a few discrete values (pasture
management in particular takes only ~5 distinct values), so treat those two as
coarse ordinal scores rather than continuous measures.

## Correlation among indices and with farm covariates

```{r}
#| label: fig-corr
#| fig-width: 7
#| fig-height: 6

bare(p_corr)
```

The four indices are only weakly-to-moderately correlated with each other, so
they carry partly independent information. Correlations with the farm covariates
(canopy cover, elevation, biomass, distance to nearest SCR farm) are modest.

## Pairwise relationships among the four indices

```{r}
#| label: fig-pairs
#| fig-width: 8
#| fig-height: 8

bare(p_pairs)
```

A closer look at the same relationships: no strong pairwise structure, and the
discreteness of pasture (and to a lesser extent water) management is visible in
the point grids.

## Representativeness of the matched farms

```{r}
#| label: fig-matched
#| fig-width: 8
#| fig-height: 6

bare(p_matched_compare)
```

The `r nrow(Farm_div_unmatched)` farms dropped for lacking a bird-diversity
estimate do not look systematically different on the diversification indices from
the `r nrow(Farm_div_matched)` retained farms, so the matched subset is
reasonably representative of the survey.

{{< pagebreak >}}

# Variation by ecoregion

## Share of variation that is between-ecoregion

```{r}
#| label: tbl-variance

Ecoregion_variance %>%
  arrange(desc(R2_ecoregion)) %>%
  kable(
    col.names = c("Type", "Variable", "n", "R² (ecoregion)", "Kruskal–Wallis p"),
    caption = "R² of `variable ~ Ecoregion`: the fraction of each variable's farm-to-farm variation that is explained by which ecoregion the farm is in."
  )
```

Bird diversity is strongly ecoregion-structured (R² $\approx$ 0.30–0.45). Among the
management indices, **water and pasture management are similarly ecoregion-bound**
(R² $\approx$ 0.34–0.35), whereas **land use and "all practices" are essentially not**
(R² $\approx$ 0.08–0.10, not significant). Elevation is almost entirely between-ecoregion
(R² $\approx$ 0.79); canopy cover is the least ecoregion-bound covariate (R² $\approx$ 0.14).

## Management diversification by ecoregion

```{r}
#| label: fig-div-bars
#| fig-width: 8
#| fig-height: 6.5

bare(p_div_bars)
```

```{r}
#| label: fig-div-box
#| fig-width: 8
#| fig-height: 6.5

bare(p_div_box)
```

Ecoregions are ordered left-to-right by ascending mean bird richness. Water and
pasture management climb along that order (lowest in Boyacá-Santander / Bajo
Magdalena, highest in Cafetera / Piedemonte); land use and "all practices" stay
roughly flat, and land use is if anything slightly *higher* in the low-diversity
ecoregions.

## Bird diversity by ecoregion (same farms)

```{r}
#| label: fig-bird-bars
#| fig-width: 8
#| fig-height: 4

bare(p_bird_bars)
```

All three Hill numbers rise across the ordered ecoregions (richness by
construction of the ordering; Shannon and Simpson track it closely).

## Farm environmental covariates by ecoregion

```{r}
#| label: fig-covar-bars
#| fig-width: 8
#| fig-height: 4

bare(p_covar_bars)
```

Median elevation and biomass differ sharply between ecoregions (elevation almost
defines them); median canopy cover overlaps much more, varying mostly within
ecoregions.

## Ecoregion signal side by side

```{r}
#| label: fig-signal
#| fig-width: 9
#| fig-height: 4.5

bare(p_std)
```

Each variable's ecoregion means, z-scored onto a common axis. The bird-diversity
lines rise smoothly; water and pasture management have a similar rising shape
(confounded with ecoregion), while land use and "all practices" do not. Elevation
and biomass peak in the high-elevation ecoregions rather than tracking the bird
gradient.

# Takeaway

At this scale, **land-use diversification and the overall "all practices" index
are the indices with usable within-ecoregion variation** — a model that accounts
for ecoregion (or the climate/topography it proxies) can still estimate an
association for those. Water and pasture management are largely a
between-ecoregion story and will be hard to separate from ecoregion itself.
