---
title: "Farm management diversification — exploratory summary and analysis plan"
subtitle: "Figures from `Scripts/02_exploratory.R`, plus notes toward the linking model"
date: today
format:
  pdf:
    toc: false
    fig-pos: "H"
    geometry: margin=1in
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

## Regenerate every figure object (p_distributions, p_corr, ... p_std) and the
## Ecoregion_variance table from the analysis script itself, so this report can
## never drift from the code. Fast -- no iNEXT bootstrap here.
source("Scripts/02_exploratory.R")

library(knitr)

## Drop the on-plot title/subtitle for the report -- the section heading and the
## paragraph under each figure carry that text, and the long subtitles clip at
## the PDF page width
bare <- function(p) p + labs(title = NULL, subtitle = NULL)
```

This report summarises the exploratory plots of the four farm management
diversification indices (land use, water management, pasture management, and an
overall "all practices" index) on the `r nrow(Farm_div_matched)` SCR farms that
have an associated bird-diversity estimate. It has three parts: (1) the indices on
their own, (2) how they — and bird diversity, and the main farm covariates — vary
among the five ecoregions, and (3) notes toward the linking model (bird diversity
as a function of the diversification indices) — how to carry the uncertainty of
the biodiversity estimates into it, and how to handle the scale of the analysis
and the confounding role of ecoregion.

# The indices on their own

## Distribution of each index

```{r}
#| label: fig-distributions
#| fig-width: 8
#| fig-height: 6

bare(p_distributions)
```

Each index runs 0–1. Land use and "all practices" are spread fairly evenly across
the range; water and pasture management pile up at a few discrete values (pasture
management in particular takes only ~5 distinct values), so treat those two as
coarse ordinal scores rather than continuous measures.

## Correlation among indices and with farm covariates

```{r}
#| label: fig-corr
#| fig-width: 7
#| fig-height: 6

bare(p_corr)
```

The four indices are only weakly-to-moderately correlated with each other, so
they carry partly independent information. Correlations with the farm covariates
(canopy cover, elevation, biomass, distance to nearest SCR farm) are modest.

## Pairwise relationships among the four indices

```{r}
#| label: fig-pairs
#| fig-width: 8
#| fig-height: 8

bare(p_pairs)
```

A closer look at the same relationships: no strong pairwise structure, and the
discreteness of pasture (and to a lesser extent water) management is visible in
the point grids.

## Representativeness of the matched farms

```{r}
#| label: fig-matched
#| fig-width: 8
#| fig-height: 6

bare(p_matched_compare)
```

The `r nrow(Farm_div_unmatched)` farms dropped for lacking a bird-diversity
estimate do not look systematically different on the diversification indices from
the `r nrow(Farm_div_matched)` retained farms, so the matched subset is
reasonably representative of the survey.

{{< pagebreak >}}

# Variation by ecoregion

## Share of variation that is between-ecoregion

```{r}
#| label: tbl-variance

Ecoregion_variance %>%
  arrange(desc(R2_ecoregion)) %>%
  kable(
    col.names = c("Type", "Variable", "n", "R² (ecoregion)", "Kruskal–Wallis p"),
    caption = "R² of `variable ~ Ecoregion`: the fraction of each variable's farm-to-farm variation that is explained by which ecoregion the farm is in."
  )
```

Bird diversity is strongly ecoregion-structured (R² $\approx$ 0.30–0.45). Among the
management indices, **water and pasture management are similarly ecoregion-bound**
(R² $\approx$ 0.34–0.35), whereas **land use and "all practices" are essentially not**
(R² $\approx$ 0.08–0.10, not significant). Elevation is almost entirely between-ecoregion
(R² $\approx$ 0.79); canopy cover is the least ecoregion-bound covariate (R² $\approx$ 0.14).

## Management diversification by ecoregion

```{r}
#| label: fig-div-bars
#| fig-width: 8
#| fig-height: 6.5

bare(p_div_bars)
```

```{r}
#| label: fig-div-box
#| fig-width: 8
#| fig-height: 6.5

bare(p_div_box)
```

Ecoregions are ordered left-to-right by ascending mean bird richness. Water and
pasture management climb along that order (lowest in Boyacá-Santander / Bajo
Magdalena, highest in Cafetera / Piedemonte); land use and "all practices" stay
roughly flat, and land use is if anything slightly *higher* in the low-diversity
ecoregions.

## Bird diversity by ecoregion (same farms)

```{r}
#| label: fig-bird-bars
#| fig-width: 8
#| fig-height: 4

bare(p_bird_bars)
```

All three Hill numbers rise across the ordered ecoregions (richness by
construction of the ordering; Shannon and Simpson track it closely).

## Farm environmental covariates by ecoregion

```{r}
#| label: fig-covar-bars
#| fig-width: 8
#| fig-height: 4

bare(p_covar_bars)
```

Median elevation and biomass differ sharply between ecoregions (elevation almost
defines them); median canopy cover overlaps much more, varying mostly within
ecoregions.

## Ecoregion signal side by side

```{r}
#| label: fig-signal
#| fig-width: 9
#| fig-height: 4.5

bare(p_std)
```

Each variable's ecoregion means, z-scored onto a common axis. The bird-diversity
lines rise smoothly; water and pasture management have a similar rising shape
(confounded with ecoregion), while land use and "all practices" do not. Elevation
and biomass peak in the high-elevation ecoregions rather than tracking the bird
gradient.

# Takeaway from the exploration

At this scale, **land-use diversification and the overall "all practices" index
are the indices with usable within-ecoregion variation** — a model that accounts
for ecoregion (or the climate/topography it proxies) can still estimate an
association for those. Water and pasture management are largely a
between-ecoregion story and will be hard to separate from ecoregion itself.

{{< pagebreak >}}

# Notes toward the linking model

## Carrying the uncertainty of the biodiversity estimates into the analysis

The response variable is itself an *estimate* — each assemblage's Hill number
comes from `02_Analysis_iNEXT.qmd` with a bootstrap standard error (`s.e.`) and a
95% interval (`qTD.LCL` / `qTD.UCL`, `nboot = 100`). Ignoring that treats a
richness estimate from one lightly-sampled farm the same as a well-sampled one.
Options, in increasing order of effort:

- **Weighted regression.** Weight each observation by `1 / s.e.^2` in
  `lmer` / `glmer` (`weights =`). Trivial to add, downweights imprecise
  assemblages — but treats the weights as known and does not propagate the
  estimation uncertainty into the coefficient intervals.

- **Bayesian measurement-error model (recommended).** In `brms`:
  `bf(y | resp_se(se, sigma = TRUE) ~ predictors + (1 | Ecoregion / Id_gcs) + (1 | CollectorXyear))`.
  Each `s.e.` enters as known measurement error while a residual $\sigma$ is
  still estimated. This folds the measurement error, the ecoregion / farm
  nesting, a non-Gaussian family, and honest posterior intervals into one
  coherent model. It is effectively a Bayesian meta-regression and fits the
  data shape well.

- **Frequentist multilevel meta-regression.**
  `metafor::rma.mv(yi = estimate, V = s.e.^2, mods = ~ predictors, random = ~ 1 | Ecoregion / Id_gcs / assemblage)`.
  Purpose-built for "effect sizes with known sampling variance", fast, handles
  the nesting. Needs a roughly Gaussian response, so log-transform the Hill
  numbers first. A good cross-check on the Bayesian model.

- **Monte-Carlo imputation of the response.** Draw B datasets, sampling each
  assemblage's diversity from its sampling distribution (Normal(estimate, `s.e.`),
  or better, resampling the retained iNEXT bootstrap replicates), refit the model
  to each, and pool with Rubin's rules. Model-agnostic — wraps around whatever
  GLMM you settle on — at the cost of more compute.

Two practical points:

- **Do not average assemblages to the farm first.** `01_Match_farm_diversity.R`'s
  `Richness_mean` etc. discard the within-farm replication and the per-assemblage
  standard errors. Model at the assemblage level with `(1 | Id_gcs)` and
  `(1 | CollectorXyear)`, and the assemblage `s.e.` as measurement error.

- Across repeated runs of `02_Analysis_iNEXT.qmd` the point estimates are
  identical but the bootstrap `s.e.` values move a little — the SE has its own
  Monte-Carlo noise at `nboot = 100`. Raise `nboot` to $\geq$ 500 before relying
  on the SEs quantitatively.

## Approach given the scale and the role of ecoregion

The central difficulty is not a modelling detail: **ecoregion is a common cause
of both bird diversity and some of the management practices**, and it explains
30–45% of the response. That is confounding plus limited overlap (water and
pasture management barely vary within a region). A workable sequence:

1. **Draw the DAG.** Ecoregion drives climate, elevation and the regional species
   pool, which drive bird diversity; ecoregion also drives baseline farm type,
   extension programmes and tenure, which drive management diversification;
   management drives habitat structure and canopy cover, which drive bird
   diversity; sampling effort drives the *observed* diversity. Ecoregion is
   almost certainly a confounder, not a mediator. Adjust for the **mechanisms it
   stands in for** (climate + elevation + a species-pool proxy) rather than the
   categorical label, so management is not competing against a factor that
   absorbs everything.

2. **Within / between (Mundlak) decomposition** — the highest-value move.
   Split each index into its ecoregion mean and the farm's deviation from that
   mean; put both in the model. The **within (deviation) coefficient** answers
   the actual question — "is a farm more diversified than its regional peers
   associated with more birds?" — and is not confounded by ecoregion-level
   differences. For water and pasture management the within coefficient will
   simply have wide intervals, which is the honest answer given how little they
   vary within region.

3. **Multilevel structure.** Assemblages within farms within ecoregions; partial
   pooling stabilises the five-region estimates: `(1 | Id_gcs)` +
   `(1 | CollectorXyear)` + `(1 | Ecoregion)`. A random slope for the index by
   ecoregion is worth a look but is likely underpowered with five regions.

4. **Be explicit about what is identifiable.** With five regions, 69 farms and
   four correlated indices you cannot cleanly estimate four separate effects.
   Either pre-register **one** primary index — land use or the "all practices"
   composite, the two with within-region variation — or run a
   **Piedemonte-only analysis** (n = 25, the most management variation, region
   held fixed) as the cleanest quasi-experiment, with the pooled five-region
   model as secondary.

5. **Model the response properly** (previous section): assemblage level,
   measurement error, an appropriate family (negative binomial for q = 0;
   Gamma or lognormal for q = 1, 2), and the sampling covariates
   (log number of point counts, number of habitats, day of year).

6. **Report variance partitioning and sensitivity, not p-values.** Use
   `partR2` / `r2glmm` for the share of variance management adds after
   ecoregion; partial-effect plots (`ggeffects`); and show the management
   coefficient across three specifications — ecoregion label, climate +
   elevation, and Mundlak. If it is stable across all three, that is the result.

7. **Causal humility.** Frame everything as association conditional on the
   adjustment set. Unmeasured farm-level confounders — farmer skill, farm size,
   baseline habitat — could plausibly drive both diversification and bird
   diversity.

**Concrete next step:** merge the indices and climate / elevation into
`03_Farm_diversity.qmd`'s assemblage-level data, then fit a `brms` model such as

```
richness | resp_se(se) ~ within_allpractices + between_allpractices +
  climate + elevation + log(Num_pc) + Num.hab +
  (1 | Id_gcs) + (1 | CollectorXyear) + (1 | Ecoregion)
```

compare it to the ecoregion-label version, and run the Piedemonte-only cut.
