---
title: "Bird diversity vs farm management diversification"
subtitle: "Bayesian measurement-error models (`Scripts/04_farm_mgmt_mod.R` + `05_farm_mgmt_plots.R`)"
date: today
format:
  pdf:
    toc: false
    number-sections: true
    fig-pos: "H"
    geometry: margin=1in
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

## Figures from the plotting script (reads the cached brms fits; no resampling).
source("Scripts/05_farm_mgmt_plots.R")

library(tidyverse)
library(knitr)

bare <- function(p) p + labs(title = NULL, subtitle = NULL)

summaries <- read_csv("Derived/Excels/Farm_mgmt_model_summaries.csv", show_col_types = FALSE)

hill_lab <- c(richness = "Richness (q = 0)", shannon = "Shannon (q = 1)", simpson = "Simpson (q = 2)")
index_lab <- c(baseline = "(baseline)", Land_use_div = "Land use", Water_mgmt_div = "Water mgmt",
               Pasture_mgmt_div = "Pasture mgmt", All_practices_div = "All practices")
spec_lab <- c(ecoregion = "Ecoregion (fixed)", climate = "Precip + elevation",
              ecoregion_numhab = "+ Num.hab (sensitivity)")
```

::: callout-note
This report is a thin wrapper around `Scripts/04_farm_mgmt_mod.R` — it renders the
figures and the summary table straight from that script's outputs, so it stays in
lock-step with the code. The narrative interpretation lives in `Project_notes.md`.
:::

# Question and design

Does **farm management diversification** — the four indices from Maria Esquivel's
spreadsheet (land use, water management, pasture management, "all practices") —
predict **bird taxonomic diversity** on the SCR farms, once region, sampling
effort and season are accounted for?

**Models** (`brms`, measurement-error via `resp_se()` on log diversity):

- **Responses:** richness (q = 0, non-asymptotic `No_Asy_TD` at Cmax + its
  coverage-based SE), Shannon (q = 1) and Simpson (q = 2) (asymptotic `TD_asy` +
  bootstrap SE).
- **Per response, a no-index _baseline_** (region + `log(Num_pc)` + habitat count
  + cyclic day of year + `(1|Id_gcs)` + `(1|CollectorXyear)`) plus one model per
  index (baseline + `index_z`).
- **Region adjusted two ways:** `Ecoregion` as a fixed effect, or farm
  precipitation + elevation. A third `ecoregion_numhab` spec is a sensitivity
  check only (habitat count partly mediates the land-use index).
- Day of year enters as `sin`/`cos` of the mean Julian day — a control for the
  Nearctic migrant influx.

# Variance explained: baseline vs adding each index

```{r}
#| label: fig-bayes-r2
#| fig-width: 9
#| fig-height: 4

bare(p_bayes_r2)
```

```{r}
#| label: tbl-r2

summaries %>%
  distinct(hill, index, spec, n_obs, bayes_R2) %>%
  filter(spec %in% c("ecoregion", "climate")) %>%
  mutate(hill = recode(hill, !!!hill_lab), index = recode(index, !!!index_lab),
         spec = recode(spec, !!!spec_lab)) %>%
  arrange(spec, hill, index) %>%
  kable(col.names = c("Response", "Index", "Region adj.", "n", "Bayesian R²"),
        caption = "Bayesian R^2 of the baseline (region + sampling + season, no index) and of each index model.")
```

# The management-diversification coefficient

```{r}
#| label: fig-index-effects
#| fig-width: 9
#| fig-height: 4

bare(p_index_effects)
```

```{r}
#| label: tbl-focal

summaries %>%
  filter(term == "focal_z", spec %in% c("ecoregion", "climate")) %>%
  mutate(cell = sprintf("%+.3f [%+.3f, %+.3f]  (P=%.2f)",
                        estimate, conf_low, conf_high, p_direction_pos)) %>%
  select(hill, index, spec, cell) %>%
  pivot_wider(names_from = spec, values_from = cell) %>%
  mutate(hill = recode(hill, !!!hill_lab), index = recode(index, !!!index_lab)) %>%
  arrange(hill, index) %>%
  kable(col.names = c("Response", "Index", "Ecoregion adj.", "Climate adj."),
        caption = "Standardized effect of a 1-SD increase in each index on log diversity: posterior median [90% CrI] and P(effect > 0).")
```

```{r}
#| label: fig-index-conditional
#| fig-width: 9
#| fig-height: 7

bare(p_index_conditional)
```

# Convergence

```{r}
#| label: tbl-diag

summaries %>%
  distinct(hill, index, spec, n_obs, max_rhat, n_divergent) %>%
  mutate(hill = recode(hill, !!!hill_lab), index = recode(index, !!!index_lab),
         spec = recode(spec, !!!spec_lab)) %>%
  arrange(desc(max_rhat)) %>%
  head(10) %>%
  kable(col.names = c("Response", "Index", "Spec", "n", "max R-hat", "divergences"),
        caption = "Ten fits with the highest R-hat (all fits shown if <= 10).")
```

```{r}
#| label: fig-ppcheck
#| fig-width: 9
#| fig-height: 8

p_ppcheck
```
