---
title: "Bird diversity, farm management diversification, and canopy-cover scale of effect"
subtitle: "`Scripts/04_farm_mgmt_mod.R`, `05_farm_mgmt_plots.R`, `06b_scale_of_effect.R`"
date: today
format:
  pdf:
    toc: true
    number-sections: true
    fig-pos: "H"
    geometry: margin=1in
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

## Rebuild the figure objects by sourcing the plotting scripts (they read cached
## fits / refit fast lmer -- no brms resampling). Keeps the report in lock-step
## with the code.
source("Scripts/05_farm_mgmt_plots.R")   # -> p_index_effects, p_bayes_r2, p_dist_sensitivity, p_ppcheck, p_index_conditional
source("Scripts/06b_scale_of_effect.R")  # -> p_scale_r2, p_scale_beta

library(tidyverse)
library(knitr)

bare <- function(p) p + labs(title = NULL, subtitle = NULL)

## fit-dependent figures are absent on an interim render (no cached brms fits)
has_fits <- !is.null(p_index_conditional)

mm_all <- read_csv("Derived/Excels/Farm_mgmt_model_summaries.csv", show_col_types = FALSE)
if (!"data_subset" %in% names(mm_all)) mm_all$data_subset <- "primary"
soe    <- read_csv("Derived/Excels/Scale_effect_results.csv", show_col_types = FALSE)

## §1.1-1.2 and §1.4 describe the primary (< 300 m) fits; the full set feeds only the §1.3 sensitivity
mm <- mm_all %>% filter(data_subset == "primary")
has_full <- "full" %in% mm_all$data_subset

subset_lab <- c(primary = "Primary (< 300 m)", full = "All assemblages")
hill_lab <- c(richness = "Richness (q = 0)", shannon = "Shannon (q = 1)", simpson = "Simpson (q = 2)")
index_lab <- c(baseline = "(baseline)", Land_use_div = "Land use", Water_mgmt_div = "Water mgmt",
               Pasture_mgmt_div = "Pasture mgmt", All_practices_div = "All practices")
spec_lab <- c(climate = "Climate + canopy (primary)", ecoregion = "Ecoregion + canopy (robustness)",
              sampling = "Sampling only", sampling_farm = "+ farm RE",
              topography = "+ topography", topography_farm = "+ topo + farm RE",
              within_eco = "Within-ecoregion (Mundlak)")
```

This repo links two independently-produced datasets on Colombia's Sustainable
Cattle Ranching farms: **bird taxonomic diversity** (iNEXT Hill numbers per
`[collector . farm . year-group . season]` assemblage,
`Scripts/00_bird_diversity_estimates.R`) and **farm management diversification**
(four `[0-1]` indices from Maria Esquivel). Two analyses: whether the
diversification indices predict bird diversity (§1), and at what spatial scale
surrounding woody-vegetation canopy cover best explains bird diversity (§2). The
assumed causal structure is in `Scripts/dag.R` (`Figures/DAG.png`).

# Bird diversity vs farm management diversification

`Scripts/04_farm_mgmt_mod.R` — `brms` measurement-error models
(`log(diversity) | resp_se(se) ~ …`) of assemblage diversity on each index, one
at a time, with a **no-index baseline** fitted alongside so the variance an index
*adds* is visible. Sampling controls throughout: `log(Num_pc)` (effort), `Num.hab`
(habitat types surveyed — a scope control, since 2013/2016-17 surveys were
forest-only and 2019+ spanned the land-use gradient), cyclic day-of-year,
`(1|Id_gcs)`, `(1|CollectorXyear)`.

Two region-adjustment versions, per the DAG:

- **Climate + canopy — primary.** Farm elevation + precipitation (each linear +
  quadratic) and surrounding 10 km canopy cover. The DAG-sufficient adjustment
  set; adjusts the mechanisms `Ecoregion` stands for without the 5-level factor
  absorbing management variation.
- **Ecoregion + canopy — robustness.** The 5-level factor as a coarse proxy for
  climate (R² ≈ 0.8) and the only handle on the regional species pool the primary
  misses; it also over-adjusts on region-correlated management variation, so it
  brackets rather than replaces the primary.

Canopy is in **both** versions — per the DAG it blocks an unobserved
farmer-values → landscape-forest backdoor, so it is a confounder to adjust, not
just a precision term.

The **primary analysis** keeps assemblages whose point counts average < 300 m
from the nominal farm (the rest sampled a different landholding); §1.3 refits on
the full set as a sensitivity. Responses: Shannon / Simpson use the asymptotic
Hill number, richness (q = 0) the coverage-standardised estimate, each with its
bootstrap / coverage-based SE.

## Variance explained: baseline vs adding each index

```{r}
#| label: fig-r2
#| fig-width: 9
#| fig-height: 4

bare(p_bayes_r2)
```

```{r}
#| label: tbl-r2

mm %>%
  distinct(hill, index, spec, n_obs, bayes_R2) %>%
  filter(spec %in% c("climate", "ecoregion")) %>%
  mutate(hill = recode(hill, !!!hill_lab), index = recode(index, !!!index_lab),
         spec = recode(spec, !!!spec_lab)) %>%
  arrange(spec, hill, index) %>%
  kable(col.names = c("Response", "Index", "Model version", "n", "Bayesian R²"),
        caption = "Bayesian R^2: baseline (region + sampling, no index) vs each index model, primary analysis.")
```

## The management-diversification coefficient

```{r}
#| label: fig-effects
#| fig-width: 9
#| fig-height: 4

bare(p_index_effects)
```

```{r}
#| label: tbl-focal

mm %>%
  filter(term == "focal_z", spec %in% c("climate", "ecoregion")) %>%
  mutate(cell = sprintf("%+.3f [%+.3f, %+.3f] (P=%.2f)", estimate, conf_low, conf_high, p_direction_pos),
         spec = recode(spec, climate = "Climate + canopy", ecoregion = "Ecoregion + canopy")) %>%
  select(hill, index, spec, cell) %>%
  pivot_wider(names_from = spec, values_from = cell) %>%
  mutate(hill = recode(hill, !!!hill_lab), index = recode(index, !!!index_lab)) %>%
  arrange(hill, index) %>%
  kable(col.names = c("Response", "Index", "Climate + canopy (primary)", "Ecoregion + canopy (robustness)"),
        caption = "Standardized effect of a 1-SD increase in each index on log diversity: posterior median [90% CrI] and P(effect > 0). Primary analysis.")
```

```{r}
#| label: fig-conditional
#| fig-width: 9
#| fig-height: 7
#| eval: !expr has_fits

bare(p_index_conditional)
```

`r if (!has_fits) "*(Conditional-effect and posterior-predictive figures omitted -- interim render, no cached model fits.)*"`

## Distance-to-farm sensitivity

The primary analysis drops assemblages whose point counts average 300 m or more
from the farm (they sampled a neighbouring landholding). This subsection refits
the index models on the **full** set and compares the focal coefficient.

```{r}
#| label: dist-sens-check
has_full <- "full" %in% mm_all$data_subset
```

`r if (!has_full) "**Interim render:** the full-set refit is not available yet; the figure/table below show the primary fits only."`

```{r}
#| label: fig-dist-sens
#| fig-width: 9
#| fig-height: 6

bare(p_dist_sensitivity)
```

```{r}
#| label: tbl-dist-sens

mm_all %>%
  filter(term == "focal_z", spec %in% c("climate", "ecoregion"), index != "baseline") %>%
  mutate(cell = sprintf("%+.3f [%+.3f, %+.3f] (n=%d)", estimate, conf_low, conf_high, n_obs)) %>%
  select(hill, index, spec, data_subset, cell) %>%
  pivot_wider(names_from = data_subset, values_from = cell) %>%
  mutate(hill = recode(hill, !!!hill_lab), index = recode(index, !!!index_lab),
         spec = recode(spec, !!!spec_lab)) %>%
  arrange(spec, hill, index) %>%
  rename_with(~ recode(.x, primary = "Primary (< 300 m)", full = "All assemblages",
                       hill = "Response", index = "Index", spec = "Model version")) %>%
  kable(caption = "Focal index coefficient (median [90% CrI], n assemblages): primary vs full set.")
```

## Diagnostics

```{r}
#| label: tbl-diag

mm_all %>%
  distinct(hill, index, spec, data_subset, n_obs, max_rhat, n_divergent) %>%
  arrange(desc(max_rhat)) %>%
  slice_head(n = 8) %>%
  mutate(hill = recode(hill, !!!hill_lab), index = recode(index, !!!index_lab),
         spec = recode(spec, !!!spec_lab), data_subset = recode(data_subset, !!!subset_lab)) %>%
  kable(col.names = c("Response", "Index", "Version", "Subset", "n", "max R-hat", "div."),
        caption = "Eight fits with the highest R-hat.")
```

```{r}
#| label: fig-ppcheck
#| fig-width: 9
#| fig-height: 8
#| eval: !expr has_fits

p_ppcheck
```

{{< pagebreak >}}

# Canopy-cover scale of effect

`Scripts/06b_scale_of_effect.R` — for each assemblage, mean woody-vegetation
canopy cover within the convex hull of its point counts buffered at 18 radii
(200 m – 10 km; `Scripts/06a_Extract_cc_buff.R`). At each radius an
`lmer(log(diversity) ~ canopy + …)` is fitted under five region-adjustment specs,
each also controlling for sampling effort, habitat count and cyclic day of year.
The "scale of effect" is the radius whose canopy term adds the most marginal R²
over its matching null. The 3 `Ref_*` reference sites are excluded.
`Figures/Scale_effect_example_buffers.png` shows one assemblage's point counts,
convex hull and rings over its canopy raster.

## Variance explained by canopy vs radius

```{r}
#| label: fig-soe-r2
#| fig-width: 11
#| fig-height: 5

bare(p_scale_r2)
```

```{r}
#| label: fig-soe-beta
#| fig-width: 11
#| fig-height: 5

bare(p_scale_beta)
```

## Best-supported radius per response and spec

```{r}
#| label: tbl-soe

soe %>%
  filter(!is.na(dR2m_canopy)) %>%
  slice_max(dR2m_canopy, n = 1, by = c(response, spec)) %>%
  transmute(
    Response = recode(str_remove(response, "_q[0-9]"), !!!hill_lab),
    Spec = recode(spec, !!!spec_lab),
    `Radius (m)` = radius_m,
    `dR2 (canopy)` = round(dR2m_canopy, 3),
    `Std. effect [95% CI]` = sprintf("%+.3f [%+.3f, %+.3f]", canopy_beta, canopy_lo, canopy_hi)
  ) %>%
  arrange(Response, Spec) %>%
  kable(caption = "Radius with the largest gain in marginal R^2 from canopy cover, per response x spec.")
```

# Takeaways

```{r}
#| label: takeaways
#| results: asis

focal <- mm %>% filter(term == "focal_z", spec %in% c("climate", "ecoregion")) %>%
  select(hill, index, spec, conf_low, conf_high) %>%
  mutate(excl0 = sign(conf_low) == sign(conf_high)) %>%
  select(hill, index, spec, excl0) %>%
  pivot_wider(names_from = spec, values_from = excl0)
n_clim <- sum(focal$climate)
n_both <- sum(focal$climate & focal$ecoregion)
best_soe <- soe %>% filter(response == "richness_q0", spec == "topography", !is.na(dR2m_canopy)) %>%
  slice_max(dR2m_canopy, n = 1)

cat("- **Management diversification:** in the primary (climate + canopy) version, ",
    if (n_clim == 0) "no index's 90% CrI excludes zero"
    else paste0(n_clim, " index-response combinations have a CrI excluding zero (pasture + water management for richness/Shannon), but ",
                n_both, " of those survive the ecoregion robustness version"),
    ". These are the two indices most confounded with region (exploratory R^2 ~ 0.35), so the attenuation under `Ecoregion` points to residual regional confounding rather than a management effect. Land use and All practices -- the region-separable indices -- are ~0 under both. Baseline and index R-squared are close throughout.\n", sep = "")
cat("- **Canopy scale of effect:** net of elevation + precipitation, bird diversity is ",
    "positively associated with surrounding canopy cover, strongest at the broad end of ",
    "the range (best radius for q = 0 richness: ", best_soe$radius_m, " m, standardized effect +",
    round(best_soe$canopy_beta, 2), "). The within-ecoregion (Mundlak) effect is weak, so ",
    "much of it is a between-region landscape gradient.\n", sep = "")
if (has_full) {
  d_p <- mm_all %>% filter(term == "focal_z", data_subset == "primary", spec %in% c("climate", "ecoregion"), index != "baseline")
  d_f <- mm_all %>% filter(term == "focal_z", data_subset == "full", spec %in% c("climate", "ecoregion"), index != "baseline")
  max_shift <- max(abs(d_p$estimate - d_f$estimate[match(
    paste(d_p$hill, d_p$index, d_p$spec), paste(d_f$hill, d_f$index, d_f$spec))]), na.rm = TRUE)
  cat("- **Distance-to-farm:** using all assemblages instead of the < 300 m primary set ",
      "shifts every index coefficient by at most ", sprintf("%.3f", max_shift),
      " SD -- the result is not driven by the cutoff (§1.3).\n", sep = "")
}
cat("- **Still to do:** a species-pool proxy (eBird) for the climate version; a ",
    "Piedemonte-only cut; LOO for the nuisance terms.\n", sep = "")
```
