---
title: "Bird diversity, farm management diversification, and canopy-cover scale of effect"
subtitle: "`Scripts/04_farm_mgmt_mod.R`, `05_farm_mgmt_plots.R`, `06b_scale_of_effect.R`"
date: today
format:
  pdf:
    toc: true
    number-sections: true
    fig-pos: "H"
    geometry: margin=1in
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

## Rebuild the figure objects by sourcing the plotting scripts (they read cached
## fits / refit fast lmer -- no brms resampling). Keeps the report in lock-step
## with the code.
source("Scripts/05_farm_mgmt_plots.R")   # -> p_index_effects, p_bayes_r2, p_dist_sensitivity, p_ppcheck, p_index_conditional
source("Scripts/06b_scale_of_effect.R")  # -> p_scale_r2, p_scale_beta

library(tidyverse)
library(knitr)

bare <- function(p) p + labs(title = NULL, subtitle = NULL)

mm_all <- read_csv("Derived/Excels/Farm_mgmt_model_summaries.csv", show_col_types = FALSE)
if (!"data_subset" %in% names(mm_all)) mm_all$data_subset <- "all"  # pre-distance-sensitivity file
soe    <- read_csv("Derived/Excels/Scale_effect_results.csv", show_col_types = FALSE)

## §2.1-2.2 and §2.4 describe the primary fits (full data); the on-farm subset feeds only the §2.3 distance-sensitivity subsection
mm <- mm_all %>% filter(data_subset == "all")

subset_lab <- c(all = "All assemblages", on_farm = "On-farm only (< 500 m)")

hill_lab <- c(richness = "Richness (q = 0)", shannon = "Shannon (q = 1)", simpson = "Simpson (q = 2)")
index_lab <- c(baseline = "(baseline)", Land_use_div = "Land use", Water_mgmt_div = "Water mgmt",
               Pasture_mgmt_div = "Pasture mgmt", All_practices_div = "All practices")
spec_lab <- c(ecoregion = "Ecoregion (fixed)", climate = "Precip + elevation",
              ecoregion_numhab = "+ Num.hab (sens.)",
              sampling = "Sampling only", sampling_farm = "+ farm RE",
              topography = "+ topography", topography_farm = "+ topo + farm RE",
              within_eco = "Within-ecoregion (Mundlak)")
```

This repo links two independently-produced datasets on Colombia's Sustainable
Cattle Ranching farms: **bird taxonomic diversity** (iNEXT Hill numbers per
`[collector . farm . year-group . season]` assemblage,
`Scripts/00_bird_diversity_estimates.R`) and **farm management diversification**
(four `[0-1]` indices from Maria Esquivel). It has two analyses: whether the
diversification indices predict bird diversity (§2), and at what spatial scale
surrounding woody-vegetation canopy cover best explains bird diversity (§3).

# Bird diversity vs farm management diversification

`Scripts/04_farm_mgmt_mod.R` — `brms` measurement-error models
(`log(diversity) | resp_se(se) ~ …`) of assemblage diversity on each index, one
at a time. Per response, a **no-index baseline** (region + `log(Num_pc)` + habitat
count + cyclic day-of-year + `(1|Id_gcs)` + `(1|CollectorXyear)`) is fitted
alongside, so the variance an index *adds* is visible. Region is adjusted two
ways — `Ecoregion` fixed, or precipitation + elevation — plus a `Num.hab`
sensitivity spec. Responses: Shannon / Simpson use the asymptotic Hill number;
richness (q = 0) uses the coverage-standardised estimate, included once its
coverage-based SE is available.

## Variance explained: baseline vs adding each index

```{r}
#| label: fig-r2
#| fig-width: 9
#| fig-height: 4

bare(p_bayes_r2)
```

```{r}
#| label: tbl-r2

mm %>%
  distinct(hill, index, spec, n_obs, bayes_R2) %>%
  filter(spec %in% c("ecoregion", "climate")) %>%
  mutate(hill = recode(hill, !!!hill_lab), index = recode(index, !!!index_lab),
         spec = recode(spec, !!!spec_lab)) %>%
  arrange(spec, hill, index) %>%
  kable(col.names = c("Response", "Index", "Region adj.", "n", "Bayesian R²"),
        caption = "Bayesian R^2: baseline (region + sampling + season, no index) vs each index model.")
```

## The management-diversification coefficient

```{r}
#| label: fig-effects
#| fig-width: 9
#| fig-height: 4

bare(p_index_effects)
```

```{r}
#| label: tbl-focal

mm %>%
  filter(term == "focal_z", spec %in% c("ecoregion", "climate")) %>%
  mutate(cell = sprintf("%+.3f [%+.3f, %+.3f] (P=%.2f)", estimate, conf_low, conf_high, p_direction_pos)) %>%
  select(hill, index, spec, cell) %>%
  pivot_wider(names_from = spec, values_from = cell) %>%
  mutate(hill = recode(hill, !!!hill_lab), index = recode(index, !!!index_lab)) %>%
  arrange(hill, index) %>%
  kable(col.names = c("Response", "Index", "Ecoregion adj.", "Climate adj."),
        caption = "Standardized effect of a 1-SD increase in each index on log diversity: posterior median [90% CrI] and P(effect > 0).")
```

```{r}
#| label: fig-conditional
#| fig-width: 9
#| fig-height: 7

bare(p_index_conditional)
```

## Distance-to-farm sensitivity

Some point counts were run off the nominal farm (`Distancia_farm` from the
wrangling repo's `Site_covs.csv`). Averaging that distance to the assemblage and
keeping only assemblages within 500 m of the farm on average, the index models
for the two primary specs are refitted on that subset and the focal coefficient
compared to the full-data fit. Point-count distances are near zero for most
assemblages, so the subset drops only a handful.

```{r}
#| label: fig-dist-sens
#| fig-width: 9
#| fig-height: 6

bare(p_dist_sensitivity)
```

```{r}
#| label: tbl-dist-sens

mm_all %>%
  filter(term == "focal_z", spec %in% c("ecoregion", "climate"), index != "baseline") %>%
  mutate(cell = sprintf("%+.3f [%+.3f, %+.3f] (n=%d)", estimate, conf_low, conf_high, n_obs)) %>%
  select(hill, index, spec, data_subset, cell) %>%
  pivot_wider(names_from = data_subset, values_from = cell) %>%
  mutate(hill = recode(hill, !!!hill_lab), index = recode(index, !!!index_lab),
         spec = recode(spec, !!!spec_lab)) %>%
  arrange(spec, hill, index) %>%
  kable(col.names = c("Response", "Index", "Region adj.", "All assemblages", "On-farm only"),
        caption = "Focal index coefficient (median [90% CrI], n assemblages): full data vs on-farm-only subset.")
```

## Diagnostics

```{r}
#| label: tbl-diag

mm_all %>%
  distinct(hill, index, spec, data_subset, n_obs, max_rhat, n_divergent) %>%
  arrange(desc(max_rhat)) %>%
  slice_head(n = 8) %>%
  mutate(hill = recode(hill, !!!hill_lab), index = recode(index, !!!index_lab),
         spec = recode(spec, !!!spec_lab), data_subset = recode(data_subset, !!!subset_lab)) %>%
  kable(col.names = c("Response", "Index", "Spec", "Subset", "n", "max R-hat", "div."),
        caption = "Eight fits with the highest R-hat.")
```

```{r}
#| label: fig-ppcheck
#| fig-width: 9
#| fig-height: 8

p_ppcheck
```

{{< pagebreak >}}

# Canopy-cover scale of effect

`Scripts/06b_scale_of_effect.R` — for each assemblage, mean woody-vegetation
canopy cover within the convex hull of its point counts buffered at 18 radii
(200 m – 10 km; `Scripts/06a_Extract_cc_buff.R`). At each radius an
`lmer(log(diversity) ~ canopy + …)` is fitted under five region-adjustment specs,
each also controlling for sampling effort, habitat count and cyclic day of year.
The "scale of effect" is the radius whose canopy term adds the most marginal R²
over its matching null. The 3 `Ref_*` reference sites are excluded.
`Figures/Scale_effect_example_buffers.png` shows one assemblage's point counts,
convex hull and rings over its canopy raster.

## Variance explained by canopy vs radius

```{r}
#| label: fig-soe-r2
#| fig-width: 11
#| fig-height: 5

bare(p_scale_r2)
```

```{r}
#| label: fig-soe-beta
#| fig-width: 11
#| fig-height: 5

bare(p_scale_beta)
```

## Best-supported radius per response and spec

```{r}
#| label: tbl-soe

soe %>%
  filter(!is.na(dR2m_canopy)) %>%
  slice_max(dR2m_canopy, n = 1, by = c(response, spec)) %>%
  transmute(
    Response = recode(str_remove(response, "_q[0-9]"), !!!hill_lab),
    Spec = recode(spec, !!!spec_lab),
    `Radius (m)` = radius_m,
    `dR2 (canopy)` = round(dR2m_canopy, 3),
    `Std. effect [95% CI]` = sprintf("%+.3f [%+.3f, %+.3f]", canopy_beta, canopy_lo, canopy_hi)
  ) %>%
  arrange(Response, Spec) %>%
  kable(caption = "Radius with the largest gain in marginal R^2 from canopy cover, per response x spec.")
```

# Takeaways

```{r}
#| label: takeaways
#| results: asis

strong <- mm %>% filter(term == "focal_z", spec == "ecoregion",
                        sign(conf_low) == sign(conf_high))
best_soe <- soe %>% filter(response == "richness_q0", spec == "topography", !is.na(dR2m_canopy)) %>%
  slice_max(dR2m_canopy, n = 1)

cat("- **Management diversification:** with `Ecoregion` adjusted for, ",
    if (nrow(strong) == 0) "no index's 90% CrI excludes zero for any response"
    else paste0(nrow(strong), " index-response combinations have a CrI excluding zero"),
    ". Baseline R-squared and index R-squared are close throughout -- the indices add little.\n", sep = "")
cat("- **Canopy scale of effect:** net of elevation + precipitation, bird diversity is ",
    "positively associated with surrounding canopy cover, strongest at the broad end of ",
    "the range (best radius for q = 0 richness: ", best_soe$radius_m, " m, standardized effect +",
    round(best_soe$canopy_beta, 2), "). The within-ecoregion (Mundlak) effect is weak, so ",
    "much of it is a between-region landscape gradient.\n", sep = "")
cat("- **Distance-to-farm:** restricting to assemblages whose point counts average ",
    "< 500 m from the farm leaves the index coefficients essentially unchanged (§2.3).\n", sep = "")
cat("- **Still to do:** the DAG; a Piedemonte-only cut; a distance-to-nearest-farm ",
    "covariate; radii beyond 10 km for a cleaner scale-of-effect peak.\n", sep = "")
```
