---
title: "Does farm management diversification benefit bird diversity?"
subtitle: "Rationale, approach, and preliminary results"
date: today
format:
  pdf:
    toc: true
    number-sections: true
    fig-pos: "H"
    geometry: margin=1in
execute:
  echo: false
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(knitr)
library(brms)

source("Scripts/05_farm_mgmt_plots.R")             # p_index_effects, p_bayes_r2
source("Scripts/05c_farm_mgmt_piedemonte_plot.R")  # p_pied_effects
source("Scripts/05d_species_pool_plot.R")          # p_env_specs, p_pool_varpart, pool_cor
bare <- function(p) p + labs(title = NULL, subtitle = NULL)

soe    <- read_csv("Derived/Excels/Scale_effect_results.csv", show_col_types = FALSE)
scheck <- read_csv("Derived/Excels/Spec_checks.csv", show_col_types = FALSE)
liksens <- read_csv("Derived/Excels/Likelihood_sensitivity.csv", show_col_types = FALSE)

hill_lab  <- c(richness = "Richness (q = 0)", shannon = "Shannon (q = 1)", simpson = "Simpson (q = 2)")
index_lab <- c(baseline = "(baseline)", Land_use_div = "Land use", Water_mgmt_div = "Water mgmt",
               Pasture_mgmt_div = "Pasture mgmt", All_practices_div = "All practices")

md <- read_csv("Derived/Excels/Farm_mgmt_model_data.csv", show_col_types = FALSE) %>% distinct(Id_gcs, .keep_all = TRUE)
idx_prec_r <- map_dbl(names(index_lab)[-1], ~ cor(md[[.x]], md$Tot_prec_mean, use = "pairwise")) %>%
  set_names(names(index_lab)[-1])
```

Source code and data-processing pipeline:
<https://github.com/LezzGitIt/Farm_diversity> (see `README.md` for the
script run order; `Scripts/dag.R` for the assumed causal structure).

# The question

The Colombian Sustainable Cattle Ranching (SCR) programme introduced
silvopasture and other land-management changes on thousands of farms.
This analysis asks the question: **do farms with more *diversified*
management have more diverse bird communities?**

It links two datasets produced independently:

- **Bird taxonomic diversity** — iNEXT Hill numbers (richness q = 0,
  Shannon q = 1, Simpson q = 2) estimated per
  `[collector . farm . year-group . season]` assemblage from
  standardised point counts (`Scripts/00_bird_diversity_estimates.R`).
- **Farm management diversification** — four `[0-1]` indices from Maria
  Esquivel: land use, water management, pasture management, and an
  overall "all practices" composite.

Joined on farm ID, `r n_distinct(md$Id_gcs)` farms have both, giving
\~`r nrow(read_csv("Derived/Excels/Farm_mgmt_model_data.csv", show_col_types = FALSE) %>% filter(Hill == "shannon"))`
assemblages.

# Why this is hard: the confounding structure

Throughout, **`Ecoregion`** means the 5-level biogeographic-region
factor (Piedemonte, Cafetera, Bajo Magdalena, Boyacá–Santander, Rio
Cesar); "region" is used only as shorthand for it.

Bird diversity here is **mostly a between-`Ecoregion` story** — the R² of
`richness ~ Ecoregion` is $\approx 0.45$. And `Ecoregion` is a
near-perfect stand-in for the environment: elevation and precipitation
are each $\approx 0.82$ explained by which ecoregion a farm sits in
(elevation on its own explains only $\approx 0.15$–$0.22$ of richness —
the diversity signal is regional, not a simple elevational gradient).
The five ecoregions differ in rainfall regime, elevation, forest cover
and biogeographic history, and those differences dominate the diversity
signal.

The four management indices **split in two** with respect to `Ecoregion`
(`Scripts/03_exploratory.R`):

- **Land use** and **All practices** are barely `Ecoregion`-structured
  (R² \~ 0.08–0.10) — they carry usable within-ecoregion variation.
- **Water management** and **Pasture management** are moderately
  `Ecoregion`-structured (R² \~ 0.34–0.35) — they co-vary with whatever
  else varies among ecoregions.

So any `Ecoregion`-correlated environmental driver of bird diversity
that also correlates with a management index is a confounder. The
assumed causal structure is in `Scripts/dag.R` (@fig-dag): `Ecoregion`
is a stand-in for climate (elevation + precipitation), landscape forest
cover, the regional species pool (via biogeographic history), and,
importantly, also for some farm management practices (particularly water
and pasture management). The identification strategy is to decompose the
`Ecoregion` category into the underlying *mechanisms* — assess the
effect of farm management while controlling for climate, landscape
forest cover, and the regional species pool.

![Assumed causal DAG behind the analysis (`Scripts/dag.R`). Two
adjustment sets are valid: climate + landscape forest + a species-pool
term + sampling, or `Ecoregion` + sampling. Farmer values, farm size and
biogeographic history stay unobserved.](/Figures/DAG.png){#fig-dag
width=100%}

The practical obstacle is **collinearity**: elevation, precipitation,
canopy cover, and the farm management practices are all highly
structured by `Ecoregion`.

# The modelling approach

`Scripts/04_farm_mgmt_mod.R` — Bayesian hierarchical regression, one
diversification index at a time:

```
log(diversity) | resp_se(se, sigma = TRUE) ~ diversification_index_z + <environment> +
    log(Num_pc)_z + Num_hab_z + doy_sin + doy_cos + (1 | farm) + (1 | collector-year)
```

- **Likelihood.** The response is not a raw count. *Observed* richness
  (q = 0) is an integer, but the model uses the iNEXT *estimate* —
  coverage-standardised richness for q = 0, asymptotic Hill-Shannon /
  Hill-Simpson for q = 1 / 2 — a continuous positive quantity read off a
  rarefaction/extrapolation curve, each with its own bootstrap standard
  error. So the likelihood is Gaussian, not a count model. It is fit on
  the **log** of the estimate for two reasons: a management or climate
  effect is naturally *multiplicative* (a proportional change in
  diversity), and iNEXT's standard error scales with the estimate, so
  `se_log = se / estimate` is roughly constant — the log scale is where
  `resp_se(se, sigma = TRUE)` behaves. It is *not* that the raw estimates
  are badly skewed: they are close to symmetric, and logging mildly
  over-corrects (@fig-loglik). The Gaussian likelihood is therefore an
  approximation on either scale; `resp_se(se, sigma = TRUE)` sets each
  assemblage's total variance to the known iNEXT error plus an estimated
  residual SD (appendix).
- **Two environmental-adjustment specifications** (`Scripts/dag.R`):
  *climate + canopy* — elevation and precipitation each as *x + x²*,
  plus surrounding 10 km canopy cover; *`Ecoregion` + canopy* — the
  5-level factor. Both are valid adjustments under the DAG (the
  `Ecoregion` specification additionally blocks a species-pool backdoor
  the climate specification leaves open; §4.2).
- Sampling controls throughout: survey effort `log(Num_pc)`, number of
  habitat types surveyed `Num_hab` (2013 / 2016-17 surveys were
  forest-only, 2019+ spanned the land-use gradient), cyclic day-of-year,
  and farm / collector-year random effects.
- A **no-index baseline** per response, so the variance an index *adds*
  is visible.
- **Primary analysis**: assemblages whose point counts average \< 300 m
  from the farm; the full set is a sensitivity analysis.

```{r}
#| label: fig-loglik
#| fig-cap: "Distribution of the assemblage diversity estimates the models are fit to. (A) Natural scale: strictly positive, only mildly right-skewed (sample skewness about 0.1 for q = 0 and q = 1, 0.4 for q = 2). (B) Log scale, with a normal matched to each response's log mean and SD (dashed): logging mildly over-corrects to a slight left skew. The two humps are the between-`Ecoregion` mixture, which the models condition on. The log scale is used for multiplicative effects and the error model, not to fix skew."
#| fig-width: 9
#| fig-height: 3.6
resp_dist <- read_csv("Derived/Excels/Farm_mgmt_model_data.csv", show_col_types = FALSE) %>%
  distinct(Assemblage, Hill, response) %>%
  filter(!is.na(response), response > 0) %>%
  mutate(q = factor(recode(Hill, !!!hill_lab), levels = unname(hill_lab)))

pal_q <- c("Richness (q = 0)" = "#1b7837", "Shannon (q = 1)" = "#762a83", "Simpson (q = 2)" = "#d95f02")

norm_ref <- resp_dist %>%
  reframe(x = seq(min(log(response)), max(log(response)), length.out = 200),
          d = dnorm(x, mean(log(response)), sd(log(response))), .by = q)

p_raw <- ggplot(resp_dist, aes(response, colour = q, fill = q)) +
  geom_density(alpha = 0.12, linewidth = 0.8) +
  scale_colour_manual(values = pal_q, name = NULL, aesthetics = c("colour", "fill")) +
  labs(x = "Diversity estimate (effective no. of species)", y = "Density", title = "(A) Natural scale") +
  cowplot::theme_minimal_grid(11)

p_log <- ggplot(resp_dist, aes(log(response))) +
  geom_density(aes(colour = q, fill = q), alpha = 0.12, linewidth = 0.8) +
  geom_line(data = norm_ref, aes(x, d, group = q), linetype = "dashed", colour = "grey35") +
  scale_colour_manual(values = pal_q, name = NULL, aesthetics = c("colour", "fill")) +
  labs(x = "log(diversity estimate)", y = "Density", title = "(B) Log scale (dashed = matched normal)") +
  cowplot::theme_minimal_grid(11)

loglik_leg <- cowplot::get_plot_component(p_raw + theme(legend.position = "bottom"),
                                          "guide-box-bottom", return_all = FALSE)
cowplot::plot_grid(
  cowplot::plot_grid(p_raw + theme(legend.position = "none"),
                     p_log + theme(legend.position = "none"), ncol = 2),
  loglik_leg, ncol = 1, rel_heights = c(1, 0.12)
)
```

Two extensions: a **Piedemonte-only** cut (`Scripts/04c_...`) that
examines diversification just within the Piedemonte ecoregion, and a
**canopy scale-of-effect** analysis (`Scripts/06b_...`).

The cyclic day-of-year term deserves a note, since the surveys span the
calendar: `doy_sin` + `doy_cos` together fit one smooth annual sine wave
of any amplitude and phase, with no discontinuity at year-end. Fitted on
the no-index Shannon baseline it is a $\approx 21\%$ peak-to-mean swing,
highest in the Sept–Oct migrant-arrival window (@fig-doy) — real enough
to be worth removing before asking about management.

```{r}
#| label: fig-doy
#| fig-cap: "Fitted seasonal (day-of-year) multiplier on diversity, no-index Shannon baseline. Shaded = Nearctic-migrant season (Sep 1 – Apr 30); ticks = the assemblages' mean survey days."
#| fig-width: 9
#| fig-height: 3.2
doy_fit <- readRDS("Derived/models/mod_shannon__baseline__climate__primary.rds")
doy_dr  <- as_draws_df(doy_fit)
doy_grid <- tibble(doy = 1:365) %>% mutate(s = sin(2 * pi * doy / 365), cc = cos(2 * pi * doy / 365))

doy_season <- map_dfr(seq_len(nrow(doy_grid)), function(i) {
  lp <- doy_dr$b_doy_sin * doy_grid$s[i] + doy_dr$b_doy_cos * doy_grid$cc[i]
  tibble(doy = doy_grid$doy[i], est = median(lp), lo = quantile(lp, 0.05), hi = quantile(lp, 0.95))
})
doy_ym     <- mean(doy_season$est)                          # centre on the annual mean -> multiplier
doy_season <- doy_season %>% mutate(across(c(est, lo, hi), ~ exp(.x - doy_ym)))
doy_days   <- read_csv("Derived/Excels/Farm_mgmt_model_data.csv", show_col_types = FALSE) %>%
  filter(Hill == "shannon") %>% distinct(Assemblage, doy)
doy_amp    <- with(doy_dr, sqrt(b_doy_sin^2 + b_doy_cos^2))
doy_mb     <- lubridate::yday(seq(as.Date("2021-01-01"), as.Date("2021-12-01"), by = "month"))

ggplot(doy_season, aes(doy, est)) +
  annotate("rect", xmin = 244, xmax = 365, ymin = -Inf, ymax = Inf, alpha = 0.08, fill = "#2166ac") +
  annotate("rect", xmin = 0, xmax = 120, ymin = -Inf, ymax = Inf, alpha = 0.08, fill = "#2166ac") +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey60") +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, fill = "#762a83") +
  geom_line(linewidth = 1, colour = "#762a83") +
  geom_rug(data = doy_days, aes(x = doy), sides = "b", alpha = 0.5, inherit.aes = FALSE) +
  scale_x_continuous(breaks = doy_mb, labels = month.abb, expand = expansion(0)) +
  labs(x = NULL, y = "Seasonal multiplier on diversity",
       subtitle = sprintf("Amplitude %.0f%% peak-to-mean (90%% CrI %.0f-%.0f%%)",
                          100 * (exp(median(doy_amp)) - 1),
                          100 * (exp(quantile(doy_amp, 0.05)) - 1),
                          100 * (exp(quantile(doy_amp, 0.95)) - 1))) +
  cowplot::theme_minimal_grid(11)
```

# Preliminary results

## No robust management effect

```{r}
#| label: fig-effects
#| fig-cap: "Standardized effect of a 1-SD increase in each diversification index on log bird diversity (posterior median, 90% CrI), primary analysis. Purple = climate + canopy model, green = Ecoregion + canopy model."
#| fig-width: 9
#| fig-height: 4
bare(p_index_effects)
```

Under the **`Ecoregion`** specification every index effect is
indistinguishable from zero (@fig-effects). Under the **climate**
specification, land use and all-practices are also $\approx 0$;
**pasture and water management are weakly positive** ($\approx +0.07$ to
$+0.08$ for richness and Shannon, credible interval sometimes excluding
zero). The baseline and index R² are close throughout — the indices add
essentially nothing (@fig-r2).

Two things temper the "no effect" reading. First, **the coefficients
are almost all positive** — the direction diversification is expected to
push diversity. Point estimates are positive in 23 of 24 index ×
response × specification combinations, and under the climate
specification $\approx 70$–$97\%$ of each posterior sits above zero
(pasture and water at the top of that range, land use and all-practices
near $70\%$; the mass is lower and closer to a coin flip under
`Ecoregion`). This is better described as a small positive effect these
data cannot resolve from zero than as evidence of no effect. Second,
**for pasture management the coefficient grows from q = 0 to q = 2** —
most clearly under the `Ecoregion` specification ($+0.035$, $+0.050$,
$+0.062$ for richness, Shannon, Simpson). Higher-q Hill numbers weight
common species more, so this is the pattern expected if diversification
benefits the common species that carry most ecosystem-service provision.
But water management shows the opposite gradient and the other two
indices are flat, so treat it as a hypothesis for the fuller dataset,
not a result.

```{r}
#| label: fig-r2
#| fig-cap: "Bayesian R² of the no-index baseline vs the model with each diversification index added (primary analysis, both model specifications). Adding an index barely moves explained variance."
#| fig-width: 9
#| fig-height: 4
bare(p_bayes_r2)
```

## The species pool is structured by elevation and precipitation

The DAG puts the regional species pool on a path from `Ecoregion` to
bird diversity (`Ecoregion` → biogeographic history → species pool → bird
diversity). The `Ecoregion` specification blocks that path; the climate
specification does not, unless the species pool is fully captured by
something in the model. So the climate specification needs a species-pool
term to be a valid adjustment.

`Scripts/01b_species_pool.R` builds one: a per-farm **potential species
pool** (`spp_pool`), the number of bird species whose Ayerbe Colombian
range polygon contains the farm **and** whose published elevational
range spans the farm's elevation. Elevational limits come mostly from
Suárez-Castro et al. (2024), with some ranges digitised manually from
field guides (Ayerbe-Quiñones 2018, Hilty) — 87 % of the 1 788 range-map
species; the rest (Nearctic migrants, seabirds) pass unfiltered.

But `spp_pool` is itself a composite of the two environmental gradients:
`spp_pool ~ poly(Elev, 2)` gives R² $\approx 0.49$,
`~ poly(precip, 2)` R² $\approx 0.52$, and both together R²
$\approx 0.87$, leaving only $\approx 13\%$ unexplained (@fig-pool-varpart).
Its own association with diversity is an **elevation** effect (strongly
positive alone, $\approx 0$ once `Elev²` is in the model), and adding it
to the management model does not move the focal coefficients
(@fig-env-specs). So this range-map count closes the species-pool
backdoor only to the extent the pool is an elevation + precipitation
composite — the *compositional* / endemism part of the pool is still
unmeasured, and an eBird- or endemism-weighted measure is the next thing
to try. In the meantime the `Ecoregion` specification is the adjustment
that closes the path cleanly.

```{r}
#| label: fig-pool-varpart
#| fig-cap: "Variance of the per-farm range-map species pool partitioned onto the elevation and precipitation gradients; together they explain about 87 percent."
#| fig-width: 9
#| fig-height: 2.6
p_pool_varpart + labs(title = NULL)
```

## The pasture / water signal is precipitation confounding

The two indices that show anything are exactly the two most correlated
with `Ecoregion` — and specifically with rainfall. At the farm level,
**`Pasture_mgmt_div` correlates with mean annual precipitation at
r = `r sprintf("%.2f", idx_prec_r["Pasture_mgmt_div"])`** (water
management r = `r sprintf("%.2f", idx_prec_r["Water_mgmt_div"])`; land
use and all-practices are slightly *negative*). Wetter farms have higher
pasture-management diversification scores, and precipitation is a
primary driver of tropical bird diversity — a confounding fork.

`Scripts/04d_species_pool_test.R` refits the model with the
environmental term specified many ways: species pool, elevation² and
precipitation² in every combination that never uses all three (they are
collinear), plus precipitation as a quadratic vs a spline, with and
without elevation co-adjusted. The pasture / water coefficient tracks
**how completely precipitation is adjusted for** (@fig-env-specs) — it
is $\approx +0.08$ to $+0.10$ whenever elevation competes with
precipitation for the shared variance (so precipitation only partly
removes the confound), and falls to $\approx +0.03$ to $+0.05$ with the
credible interval spanning zero whenever precipitation is the sole or
dominant environmental term. Quadratic vs spline makes no difference.

```{r}
#| label: fig-env-specs
#| fig-cap: "Management-index coefficient (pasture, water) across every environmental specification, labelled by the variables it contains and ordered by how completely precipitation is controlled. 10 km canopy and the sampling controls are held fixed throughout. The coefficient shrinks toward zero as precipitation adjustment improves."
#| fig-width: 9
#| fig-height: 6
bare(p_env_specs)
```

## Holding `Ecoregion` fixed: Piedemonte only

```{r}
#| label: fig-pied
#| fig-cap: "Piedemonte-only fits (`Scripts/04c`): standardized management-index effect on log diversity, `Ecoregion` held fixed, elevation/precipitation/canopy replaced by their first PC. Primary (< 300 m) vs all assemblages. Every credible interval includes zero."
#| fig-width: 9
#| fig-height: 4
bare(p_pied_effects)
```

Restricting to the 25 Piedemonte farms (`Scripts/04c_...`), with the
collinear elevation / precipitation / canopy replaced by their first
principal component, still shows **no convincing management effect**.
This is consistent with the confounding story: management is generally
positive, especially pasture / water management, but every credible
interval includes zero.

## A landscape-context effect: canopy cover scale of effect

```{r}
#| label: tbl-soe
soe %>%
  filter(!is.na(dR2m_canopy), spec %in% c("topography", "topography_farm", "within_eco")) %>%
  slice_max(dR2m_canopy, n = 1, by = c(response, spec)) %>%
  transmute(
    Response = recode(str_remove(response, "_q[0-9]"), !!!hill_lab),
    Spec = recode(spec, topography = "+ elevation + precip", topography_farm = "+ topo + farm RE",
                  within_eco = "within-ecoregion (Mundlak)"),
    `Radius (m)` = radius_m,
    `Std. effect [95% CI]` = sprintf("%+.3f [%+.3f, %+.3f]", canopy_beta, canopy_lo, canopy_hi)
  ) %>%
  arrange(Response, Spec) %>%
  kable(caption = "Canopy-cover scale of effect (Scripts/06b): net of elevation + precipitation, bird richness is positively associated with surrounding woody-vegetation cover, strongest at a broad ~8-9 km radius. This is a landscape-context effect, not a farm-management one.")
```

## Specification robustness

Four modelling choices were checked (`Scripts/04e_spec_checks.R`); none
changes the conclusion.

**`(1 | CollectorXyear)`** — the \[dataset × year-group\] random effect
(8 levels) is only partly confounded with `Ecoregion` (Cramér's V =
0.38; the three largest datasets each span 4–5 ecoregions). Its standard
deviation is 0.18–0.31 on the log scale, so it captures real
field-method differences among the six source datasets. Dropping it
shifts every management coefficient by $\le 0.02$, on the full data and
in Piedemonte alike — it is not standing in for the `Ecoregion` factor.

**`resp_se()`** — propagating the iNEXT standard errors makes the fit a
precision-weighted regression. Dropping it (treating the point estimates
as exact) gives \~25 % *larger* management coefficients and \~7 % wider
credible intervals, because the well-sampled assemblages — which show a
weaker signal — then count for less. `resp_se` is both the honest and
the conservative choice.

**Distance-to-farm cutoff** — for some assemblages the point counts
average well off the nominal farm (@fig-dist): different sampling teams
placed counts along access roads or on neighbouring land. The primary
analysis keeps only assemblages whose counts average \< 300 m from the
farm; using all assemblages instead shifts the coefficients by
$\le 0.03$, toward zero if anything.

**Response scale / likelihood** — the model is Gaussian on `log(diversity)`,
but the raw estimates are close to symmetric (@fig-loglik), so is the
result an artefact of the log scale? Refitting the pasture and water ×
richness / Shannon models as Gaussian-on-log, Student-t-on-log (robust to
the left tail) and Gaussian-on-raw — all with `resp_se`, the only
alternatives that support it — and expressing the effect as a percent
change in diversity per 1-SD increase in the index gives the same answer
each way (@tbl-liksens): a small positive effect, credible interval
usually spanning zero.

```{r}
#| label: tbl-liksens
liksens %>%
  select(index, hill, likelihood, pct_effect) %>%
  pivot_wider(names_from = likelihood, values_from = pct_effect) %>%
  arrange(index, hill) %>%
  kable(caption = "Focal effect as a percent change in diversity per +1 SD of the index (climate specification, < 300 m), under three likelihoods that all support resp_se.")
```

```{r}
#| label: fig-dist
#| fig-cap: "Mean distance of an assemblage's point counts from its farm. Dashed line = the 300 m primary-analysis cutoff; the long right tail is dropped from the primary set."
#| fig-width: 9
#| fig-height: 2.8
read_csv("Derived/Excels/Farm_mgmt_model_data.csv", show_col_types = FALSE) %>%
  filter(Hill == "shannon", !is.na(dist_farm)) %>%
  distinct(Assemblage, dist_farm) %>%
  ggplot(aes(dist_farm)) +
  geom_histogram(binwidth = 100, boundary = 0, fill = "grey70", colour = "white") +
  geom_vline(xintercept = 300, linetype = "dashed", colour = "#d95f02") +
  labs(x = "Mean distance of the point counts from the farm (m)", y = "Assemblages") +
  theme_minimal(11)
```

```{r}
#| label: tbl-scheck
scheck %>%
  filter(env == "climate") %>%
  mutate(variant = recode(variant, base = "base", no_collector = "no CollectorXyear RE",
                          no_resp_se = "no resp_se"),
         hill = recode(hill, !!!hill_lab), index = recode(index, !!!index_lab)) %>%
  select(hill, index, variant, focal_est) %>%
  pivot_wider(names_from = variant, values_from = focal_est) %>%
  arrange(hill, index) %>%
  kable(digits = 3,
        caption = "Focal coefficient (posterior median) under the climate specification: base vs dropping the CollectorXyear random effect vs dropping resp_se. Primary analysis.")
```

# Interpretation

*These are preliminary results on a partial dataset (69 farms, several
indices incomplete). The conclusions below are what the current data
support, not settled findings.*

**At this scale, with these data, farm management diversification has no
clearly detectable effect on bird taxonomic diversity** — though the
coefficients lean positive throughout (§4.1).

- The **land-use** and **all-practices** indices — the ones separable
  from `Ecoregion` — are near-null under every specification, with
  posteriors only mildly favouring a positive effect.
- The **pasture** and **water-management** indices show a weak positive
  coefficient under the parametric climate adjustment, but it tracks how
  well precipitation is controlled: it is $\approx +0.08$ when
  precipitation is weakly adjusted and $\approx +0.04$ with the credible
  interval spanning zero once precipitation is adjusted well — including
  under the `Ecoregion` specification, which is R² 0.83 with
  precipitation. So at least part of it is **precipitation confounding**.
- The environmental gradients dominate, and the `Ecoregion`-bound
  indices are too collinear with them for a management effect to be
  cleanly identified from observational data.
- Separately, **surrounding landscape forest cover** (\~8–9 km) *is*
  positively associated with bird richness, net of climate — a
  landscape-context effect that is not about on-farm management.

## Why `Ecoregion` is legitimate adjustment, not over-adjustment

The concern with the 5-level factor was that it might absorb
`Ecoregion`–management covariation that is not confounding — e.g. if the
SCR programme promoted water-management practices more in wetter
ecoregions as policy, conditioning on `Ecoregion` would regress away
part of a real effect. Three things argue against that:

1.  **The attenuation is reproducible with no `Ecoregion` factor at
    all**, purely by making precipitation the dominant environmental
    term (dropping the elevation term from the continuous model).
    Continuous climate variables do not carry a "the factor soaks up
    management structure" mechanism, so if the `Ecoregion` result were
    over-adjustment they could not reproduce it — but they do.
2.  **There is a concrete confounding path**:
    `Pasture_mgmt_div ~ precipitation` r = 0.30 (and only 0.02 with
    elevation), and precipitation drives bird diversity. The DAG has
    this fork.
3.  **The `climate` specification controls precipitation *worse* than
    the factor, not better.** Elevation and precipitation go in as two
    negatively-correlated continuous terms; because they trade off,
    precipitation removes less of the pasture / water confound than it
    would alone. Drop the elevation term — or use the single flexible
    `Ecoregion` factor — and precipitation absorbs more of the confound:
    the coefficient falls to \~0.04 with the credible interval spanning
    zero, regardless of whether precipitation enters as a quadratic or a
    spline. So the `climate`–`Ecoregion` gap is the parametric climate
    specification *under-adjusting* precipitation under collinearity, not
    the factor over-adjusting.

**Caveat.** With everything this collinear, "legitimate adjustment" is
an inference from the pattern, not a proof. Precipitation and regional
programme rollout may themselves be confounded, so we cannot fully
separate "wetter, so more birds" from "wetter ecoregion, so more
water-management promotion". But under either reading the
well-precipitation-adjusted estimate — small and not distinguishable
from zero — is the one to report, not the $+0.08$.

# Where this goes

- **A better environmental term in the `climate` specification** — the
  parametric `Elev + Elev² + precip + precip²` under-controls
  precipitation because the two collinear axes trade off (§5.1). A single
  flexible term (an `s(lat, long)` spatial spline, or dropping the
  elevation quadratic), or simply taking the `Ecoregion` result as
  primary, would close the `climate`–`Ecoregion` gap at \~0.
- **Other response axes** — functional or phylogenetic diversity, or
  species-of-conservation-concern subsets, may respond to management
  where taxonomic Hill numbers do not.
- **The biogeographic residual** — an endemism-weighted pool or an eBird
  composition measure, if the compositional part of the `Ecoregion`
  effect is worth chasing.
- **Farm isolation** — `dist_predio_cercano` (distance to the nearest
  neighbouring property, from the MJE spreadsheet) is uncorrelated with
  the point-count `Distancia_farm` ($r \approx 0$), so it is a genuinely
  new covariate: a candidate predictor and/or a spatial-autocorrelation
  control.
- **Manuscript framing** — the honest story is "farm-scale management
  diversification does not detectably raise bird taxonomic diversity
  once the environmental gradient is accounted for, though the effect is
  consistently positive; surrounding landscape forest cover does" —
  reported with the collinearity, sample-size and observational caveats.
- Better understanding of iNEXT diversification indices - Consideration
  of alternative cutoff values.

# Appendix: methodological questions

Six questions that came up while building the analysis, with the answers
that shaped the current specification.

## Is canopy cover "required by the DAG"?

Not as a farmer-values backdoor — that was a scale mismatch. The canopy
covariate is woody cover in a $\approx 10$ km buffer, set by topography,
protected areas and hundreds of landholders; no single farmer's
priorities move it. So there is no farmer-values-to-canopy arrow and
canopy is not blocking a farmer-values path. What canopy legitimately
does: (1) in the `climate` specification it blocks the `Ecoregion`
backdoor FarmDiv \<- Ecoregion -\> LandForest -\> BirdDiv; (2)
everywhere, it is a strong predictor of diversity, so it reduces
residual variance. Under the `Ecoregion` specification it is not needed
at all but is harmless and variance-reducing, so it stays in both. The
genuine residual concern —
bird-minded farmers doing *on-farm* things the four indices miss (snag
retention, less pesticide, wider fencerows) — is a farm-scale path that
canopy cannot proxy; it is left as an explicit unmeasured-confounder
caveat in `Scripts/dag.R`.

## Why is farm size in the DAG?

Farm area (unmeasured) has three arrows. FarmSize -\> FarmDiv: a bigger
farm has more room and more management units, so more opportunity to
score high on indices that are partly counts of practices.
FarmSize -\> NumPC: bigger farms get more point counts.
FarmSize -\> BirdDiv: species–area — more area means larger populations,
more interior habitat, proportionally less edge. Conditioning on
`NumPC` (already in every model) closes the
FarmDiv \<- FarmSize -\> NumPC -\> BirdDiv path — this is why `NumPC` is
now a genuine confounder in the DAG's minimal adjustment set, not just a
precision term. The direct FarmSize -\> BirdDiv path is **not** closed
by anything in the model and would bias a true null upward. Measuring
farm hectares would close it; it is on the re-analysis checklist.

## `resp_se(se, sigma = TRUE)` vs `sigma = FALSE`

With `sigma = TRUE` the total variance for assemblage *i* is
$se_i^2 + \sigma^2$ — the known iNEXT bootstrap SE **plus** an estimated
residual SD shared across rows. `sigma = FALSE` uses only $se_i^2$
(pure inverse-variance / meta-analysis weighting), which assumes the
known SEs are the whole error story. Refitting pasture × richness
(climate primary) both ways: `TRUE` gives
$+0.081\ [-0.002,\ 0.164]$, `FALSE` gives
$+0.088\ [-0.007,\ 0.184]$ (a slightly larger point estimate and a
$\approx 15\%$ wider interval). The `TRUE` model's estimated residual SD
is $\approx 0.15$, larger than the mean known measurement error
($\overline{se_{\log}} \approx 0.09$) — there is real
assemblage-to-assemblage scatter beyond the bootstrap SE, so
`sigma = TRUE` is the correct choice. (This is separate from the
"drop `resp_se` entirely" check in §4.6, which also discards the known
SEs and gives $\approx 25\%$ larger coefficients.)

## Removing the $\pm 250$ m species-pool elevation buffer

Done. A farm's elevation must now fall inside a species' published
elevational range exactly. This keeps 96 % of the range-polygon hits
(was 98 %); `spp_pool ~ climate` R² and the management coefficients in
@fig-env-specs are unchanged.

## What does it mean if Bayesian R² *decreases* when an index is added?

Almost entirely a subset artifact, not a sign the index hurts. Bayesian
R² is computed on the rows each model was fit on. The no-index baseline
uses all assemblages with complete covariates ($n \approx 89$–$98$); the
water- and pasture-management models drop farms missing those indices
($n \approx 73$–$80$). Different data, so the R² values are not directly
comparable — removing $\approx 20$ rows changes both the fitted-value
variance and the residual variance. Two smaller contributors: the
`normal(0, 0.75)` prior shrinks a near-zero coefficient toward 0, so an
uninformative index adds almost nothing to explained variance; and for
Shannon/Simpson the water/pasture *subset* actually has a **higher** R²
than the baseline, only richness is slightly lower — that inconsistency
across responses is the signature of subset noise. A clean comparison
would refit the baseline on the matched $n$ (backlog).

## Would LOO for the nuisance terms change anything now?

Not at this stage. With $n \approx 73$–$98$ and the regularising
`normal(0, 0.75)` prior, `Scripts/04e_spec_checks.R` already shows that
dropping a nuisance block (the CollectorXyear RE, and by extension the
day-of-year terms) moves the focal coefficient by $\le 0.02$. LOO would
formalise "keep them" without changing the conclusion, and LOO with this
many single-assemblage farms and the measurement-error structure needs
care (Pareto-$k$ diagnostics, possibly moment-matching). It is worth
doing when the specifications are being finalised for the manuscript on
a larger dataset, not now.
