---
title: "Does farm management diversification benefit bird diversity?"
subtitle: "Rationale, approach, and preliminary results"
date: today
format:
  pdf:
    toc: true
    number-sections: true
    fig-pos: "H"
    geometry: margin=1in
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(knitr)

source("Scripts/05_farm_mgmt_plots.R")        # p_index_effects, p_bayes_r2
source("Scripts/05c_farm_mgmt_piedemonte_plot.R")  # p_pied_effects
source("Scripts/05d_species_pool_plot.R")     # p_pool_forest, p_pool_varpart, pool_cor
bare <- function(p) p + labs(title = NULL, subtitle = NULL)

mm   <- read_csv("Derived/Excels/Farm_mgmt_model_summaries.csv", show_col_types = FALSE) %>% filter(data_subset == "primary")
soe  <- read_csv("Derived/Excels/Scale_effect_results.csv", show_col_types = FALSE)
sens <- read_csv("Derived/Excels/Precip_vs_elev_sensitivity.csv", show_col_types = FALSE)

hill_lab  <- c(richness = "Richness (q = 0)", shannon = "Shannon (q = 1)", simpson = "Simpson (q = 2)")
index_lab <- c(baseline = "(baseline)", Land_use_div = "Land use", Water_mgmt_div = "Water mgmt",
               Pasture_mgmt_div = "Pasture mgmt", All_practices_div = "All practices")

md <- read_csv("Derived/Excels/Farm_mgmt_model_data.csv", show_col_types = FALSE) %>% distinct(Id_gcs, .keep_all = TRUE)
idx_prec_r <- map_dbl(names(index_lab)[-1], ~ cor(md[[.x]], md$Tot_prec_mean, use = "pairwise")) %>%
  set_names(names(index_lab)[-1])
```

# The question

The Colombian Sustainable Cattle Ranching (SCR) programme introduced silvopasture
and other land-management changes on thousands of farms. This analysis asks a
narrower question: **do farms with more *diversified* management have more diverse
bird communities?**

It links two datasets produced independently:

- **Bird taxonomic diversity** — iNEXT Hill numbers (richness q = 0, Shannon
  q = 1, Simpson q = 2) estimated per `[collector . farm . year-group . season]`
  assemblage from standardised point counts (`Scripts/00_bird_diversity_estimates.R`).
- **Farm management diversification** — four `[0-1]` indices from Maria Esquivel:
  land use, water management, pasture management, and an overall "all practices"
  composite.

Joined on farm ID, `r n_distinct(md$Id_gcs)` farms have both, giving
~`r nrow(read_csv("Derived/Excels/Farm_mgmt_model_data.csv", show_col_types = FALSE) %>% filter(Hill == "shannon"))`
assemblages.

# Why this is hard: the confounding structure

Bird diversity in this region is **mostly a between-ecoregion story** — the R²
of `richness ~ Ecoregion` is ≈ 0.45, and elevation alone is ≈ 0.79. The five
ecoregions differ in rainfall regime, elevation, forest cover and biogeographic
history, and those differences dominate the diversity signal.

The four management indices **split in two** with respect to region
(`Scripts/03_exploratory.R`):

- **Land use** and **All practices** are barely region-structured
  (R² ~ 0.08–0.10) — they carry usable within-region variation.
- **Water management** and **Pasture management** are strongly region-structured
  (R² ~ 0.34–0.35) — they co-vary with whatever else varies by region.

So any region-correlated environmental driver of bird diversity that also
correlates with a management index is a **confounder**. The assumed causal
structure is in `Scripts/dag.R` (Figure below): `Ecoregion` is a stand-in for
**climate (elevation + precipitation), landscape forest cover, and the regional
species pool** — and the identification strategy is to adjust *those mechanisms*
rather than the categorical label, so the label does not also soak up
non-confounding region–management structure.

The practical obstacle is **collinearity**: elevation, precipitation, canopy
cover, ecoregion and the region-bound indices are all entangled, so the
mechanisms cannot be separated cleanly.

# The modelling approach

`Scripts/04_farm_mgmt_mod.R` — Bayesian hierarchical regression, one index at a
time:

```
log(diversity) | resp_se(se, sigma = TRUE) ~ index_z + <environment> +
    log(Num_pc)_z + Num_hab_z + doy_sin + doy_cos + (1 | farm) + (1 | collector-year)
```

- `resp_se()` propagates each iNEXT estimate's bootstrap / coverage-based
  standard error into the response.
- **Two environmental-adjustment versions** (`Scripts/dag.R`):
  *climate + canopy* — elevation and precipitation each as *x + x²*, plus
  surrounding 10 km canopy cover (the DAG-sufficient set, primary);
  *Ecoregion + canopy* — the 5-level factor as a coarse proxy (robustness check).
- Sampling controls throughout: survey effort `log(Num_pc)`, number of habitat
  types surveyed `Num_hab` (2013 / 2016-17 surveys were forest-only, 2019+
  spanned the land-use gradient), cyclic day-of-year, and farm / collector-year
  random effects.
- A **no-index baseline** per response, so the variance an index *adds* is
  visible.
- **Primary analysis**: assemblages whose point counts average < 300 m from the
  farm (the rest sampled a neighbouring landholding); the full set is a
  sensitivity.

Two extensions: a **Piedemonte-only** cut (`Scripts/04c_...`) that holds region
fixed, and a **canopy scale-of-effect** analysis (`Scripts/06b_...`).

# Preliminary results

## No robust management effect

```{r}
#| label: fig-effects
#| fig-width: 9
#| fig-height: 4
bare(p_index_effects)
```

```{r}
#| label: tbl-focal
mm %>%
  filter(term == "focal_z", spec %in% c("climate", "ecoregion")) %>%
  mutate(cell = sprintf("%+.3f [%+.3f, %+.3f]", estimate, conf_low, conf_high),
         spec = recode(spec, climate = "Climate + canopy", ecoregion = "Ecoregion + canopy"),
         hill = recode(hill, !!!hill_lab), index = recode(index, !!!index_lab)) %>%
  select(hill, index, spec, cell) %>%
  pivot_wider(names_from = spec, values_from = cell) %>%
  arrange(hill, index) %>%
  kable(caption = "Standardized effect of a 1-SD increase in each index on log diversity: posterior median [90% CrI], primary analysis.")
```

Under the **Ecoregion** version every index effect is indistinguishable from
zero. Under the **climate** version, land use and all-practices are also ≈ 0;
**pasture and water management are weakly positive** (~+0.07 to +0.08 for
richness and Shannon, credible interval sometimes excluding zero). The baseline
and index R² are close throughout — the indices add essentially nothing
(@fig-r2).

```{r}
#| label: fig-r2
#| fig-width: 9
#| fig-height: 4
bare(p_bayes_r2)
```

## The pasture / water signal is precipitation confounding

The two indices that show anything are exactly the two most correlated with
region — and specifically with rainfall. At the farm level,
**`Pasture_mgmt_div` correlates with mean annual precipitation at r =
`r sprintf("%.2f", idx_prec_r["Pasture_mgmt_div"])`** (water management r =
`r sprintf("%.2f", idx_prec_r["Water_mgmt_div"])`; land use and all-practices are
slightly *negative*). Wetter farms diversify pasture management more, and
precipitation is a primary driver of tropical bird diversity — a textbook
confounding fork.

`Scripts/04d_species_pool_test.R` and a follow-up sensitivity fit the model with
the environmental term specified several ways. The pasture / water coefficient
tracks **how hard precipitation is controlled**, nothing else:

```{r}
#| label: tbl-sens
sens %>%
  select(index, hill, block, focal) %>%
  kable(col.names = c("Index", "Response", "Environmental block", "Focal coefficient"),
        caption = "Focal coefficient (median [95% CrI]) with the environmental term specified four ways. It is ~+0.08 whenever elevation is in and precipitation only weakly controlled, and ~+0.03-0.04 (CrI includes 0) whenever precipitation is the dominant control. Spline vs quadratic makes no difference.")
```

```{r}
#| label: fig-pool-forest
#| fig-width: 9
#| fig-height: 6
bare(p_pool_forest)
```

*Reading @fig-pool-forest:* four environmental blocks, none combining all
three of `spp_pool`, `Elev²`, `precip²` (they are collinear). The management
coefficient (green) is largest in the block where precipitation is least
controlled (`pool_elev`) and smallest where it is most controlled
(`pool_precip`). `spp_pool` itself (purple) is defined below.

## The species pool does not disentangle it

`Scripts/01b_species_pool.R` builds a per-farm **potential species pool**
(`spp_pool`): the number of bird species whose Ayerbe Colombian range polygon
contains the farm **and** whose elevational limits span the farm's elevation
± 250 m. Elevational limits come from Suárez-Castro et al. (2024) backfilled with
the manual field-guide digitisations in `03_FT_elev.R` (Ayerbe-Quiñones 2018,
Hilty) — 87 % of the 1 788 range-map species; the rest (Nearctic migrants,
seabirds) pass unfiltered. This is external to the point-count data, so it is a
genuine "how many species could occur here", not a circular re-measurement.

But `spp_pool` is itself a composite of the two environmental gradients —
`spp_pool ~ poly(Elev, 2)` gives R² ≈ 0.43, `~ poly(precip, 2)` R² ≈ 0.63, and
both together R² ≈ 0.89, leaving only ~11 % unexplained:

```{r}
#| label: fig-pool-varpart
#| fig-width: 9
#| fig-height: 2.6
p_pool_varpart + labs(title = NULL)
```

```{r}
#| label: tbl-pool-cor
pool_cor %>% kable(col.names = c("Gradient", "correlation with spp_pool"),
                   caption = "spp_pool tracks precipitation positively and elevation negatively (montane depauperation).")
```

Adding `spp_pool` to the model does not move the management coefficients
(@fig-pool-forest, green points barely shift between `clim` and `pool`),
and `spp_pool`'s own association with diversity is an **elevation** effect
(strongly positive alone, ~0 once `Elev²` is in). So a range-map species-pool
count adds no information beyond elevation + precipitation. A composition- or
endemism-weighted pool, or an eBird-based measure, would be the next thing to try
if the aim is a *biogeographic* residual.

## Holding region fixed: Piedemonte only

```{r}
#| label: fig-pied
#| fig-width: 9
#| fig-height: 4
bare(p_pied_effects)
```

Restricting to the 25 Piedemonte farms (`Scripts/04c_...`), with the collinear
elevation / precipitation / canopy replaced by their first principal component,
still shows **no detectable management effect** — every credible interval
includes zero. This is consistent with the confounding story rather than an
independent confirmation: the principal component is only about one-third
precipitation (it weights elevation, precipitation and canopy equally), so
precipitation is weakly controlled here, and pasture / water sit at the same
~+0.06 as the weak-precipitation-control full models.

## A landscape-context effect: canopy cover scale of effect

```{r}
#| label: tbl-soe
soe %>%
  filter(!is.na(dR2m_canopy), spec %in% c("topography", "topography_farm", "within_eco")) %>%
  slice_max(dR2m_canopy, n = 1, by = c(response, spec)) %>%
  transmute(
    Response = recode(str_remove(response, "_q[0-9]"), !!!hill_lab),
    Spec = recode(spec, topography = "+ elevation + precip", topography_farm = "+ topo + farm RE",
                  within_eco = "within-ecoregion (Mundlak)"),
    `Radius (m)` = radius_m,
    `Std. effect [95% CI]` = sprintf("%+.3f [%+.3f, %+.3f]", canopy_beta, canopy_lo, canopy_hi)
  ) %>%
  arrange(Response, Spec) %>%
  kable(caption = "Canopy-cover scale of effect (Scripts/06b): net of elevation + precipitation, bird richness is positively associated with surrounding woody-vegetation cover, strongest at a broad ~8-9 km radius. This is a landscape-context effect, not a farm-management one.")
```

# Interpretation

**At this scale, with these data, farm management diversification shows no
detectable effect on bird taxonomic diversity.**

- The **land-use** and **all-practices** indices — the ones separable from region
  — are flat null under every specification.
- The **pasture** and **water-management** indices show a weak positive
  coefficient under the parametric climate adjustment, but it is a
  **precipitation-confounding artifact**: it appears only when precipitation is
  weakly controlled and disappears (credible interval spanning zero) whenever
  precipitation is adjusted well — including under the `Ecoregion` factor, which
  is R² 0.83 with precipitation.
- The environmental gradients dominate, and the region-bound indices are too
  collinear with them for a management effect to be identified from observational
  data.
- Separately, **surrounding landscape forest cover** (~8–9 km) *is* positively
  associated with bird richness, net of climate — a landscape-context effect that
  is not about on-farm management.

## Why `Ecoregion` is legitimate adjustment, not over-adjustment

The concern with the 5-level factor was that it might absorb region–management
covariation that is not confounding — e.g. if the SCR programme promoted
water-management practices more in wetter regions as policy, conditioning on
`Ecoregion` would regress away part of a real effect. Three things argue against
that:

1. **The attenuation is reproducible with no region factor at all**, purely by
   controlling precipitation better (a precipitation spline, or a spec where
   precipitation dominates the adjustment). Continuous climate variables do not
   carry a "the factor soaks up management structure" mechanism, so if the
   `Ecoregion` result were over-adjustment they could not reproduce it — but they
   do.
2. **There is a concrete confounding path**: `Pasture_mgmt_div ~ precipitation`
   r = 0.30, and precipitation drives bird diversity. The DAG has this fork.
3. **The `climate` spec under-controls precipitation relative to the factor** —
   `precip + precip²` is a rigid global curve; the 5-level factor is a flexible
   step (R² 0.83 with precipitation). Giving the continuous spec a precipitation
   spline moves it toward the factor's near-zero answer. So the gap is a
   *precision-of-adjustment* gap, not an over- vs correct-adjustment gap.

**Caveat.** With everything this collinear, "legitimate adjustment" is an
inference from the pattern, not a proof. Precipitation and "regional programme
rollout" may themselves be confounded, so we cannot fully separate "wetter → more
birds" from "wetter region → more water-management promotion". But under either
reading the direction is the same: the well-precipitation-adjusted near-zero
estimate is the one to report, and the +0.08 is not a defensible management
effect.

# Where this goes

- **DAG-consistent precipitation term** — replace `precip + precip²` in the
  climate spec with `s(precip)` so it is controlled as well as the `Ecoregion`
  factor does it; the two specs should then converge at ~0.
- **Other response axes** — functional or phylogenetic diversity, or
  species-of-conservation-concern subsets, may respond to management where
  taxonomic Hill numbers do not.
- **The biogeographic residual** — an endemism-weighted pool or an eBird
  composition measure, if the compositional part of the `Ecoregion` effect is
  worth chasing.
- **Manuscript framing** — the honest story is "farm-scale management
  diversification does not detectably raise bird taxonomic diversity once the
  environmental gradient is accounted for; surrounding landscape forest cover
  does" — reported with the collinearity and observational caveats.
