---
title: "Does farm management diversification benefit bird diversity?"
subtitle: "Rationale, approach, and preliminary results"
date: today
format:
  pdf:
    toc: true
    number-sections: true
    fig-pos: "H"
    geometry: margin=1in
execute:
  echo: false
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(knitr)

source("Scripts/05_farm_mgmt_plots.R")             # p_index_effects, p_bayes_r2
source("Scripts/05c_farm_mgmt_piedemonte_plot.R")  # p_pied_effects
source("Scripts/05d_species_pool_plot.R")          # p_env_specs, p_pool_varpart, pool_cor
source("Scripts/05e_doy_explainer.R")              # p_fit (fitted seasonal effect)
bare <- function(p) p + labs(title = NULL, subtitle = NULL)

soe    <- read_csv("Derived/Excels/Scale_effect_results.csv", show_col_types = FALSE)
scheck <- read_csv("Derived/Excels/Spec_checks.csv", show_col_types = FALSE)

hill_lab  <- c(richness = "Richness (q = 0)", shannon = "Shannon (q = 1)", simpson = "Simpson (q = 2)")
index_lab <- c(baseline = "(baseline)", Land_use_div = "Land use", Water_mgmt_div = "Water mgmt",
               Pasture_mgmt_div = "Pasture mgmt", All_practices_div = "All practices")

md <- read_csv("Derived/Excels/Farm_mgmt_model_data.csv", show_col_types = FALSE) %>% distinct(Id_gcs, .keep_all = TRUE)
idx_prec_r <- map_dbl(names(index_lab)[-1], ~ cor(md[[.x]], md$Tot_prec_mean, use = "pairwise")) %>%
  set_names(names(index_lab)[-1])
```

Source code and data-processing pipeline:
<https://github.com/LezzGitIt/Farm_diversity> (see `README.md` for the
script run order; `Scripts/dag.R` for the assumed causal structure).

# The question

The Colombian Sustainable Cattle Ranching (SCR) programme introduced
silvopasture and other land-management changes on thousands of farms.
This analysis asks the question: **do farms with more *diversified*
management have more diverse bird communities?**

It links two datasets produced independently:

- **Bird taxonomic diversity** — iNEXT Hill numbers (richness q = 0,
  Shannon q = 1, Simpson q = 2) estimated per
  `[collector . farm . year-group . season]` assemblage from
  standardised point counts (`Scripts/00_bird_diversity_estimates.R`).
- **Farm management diversification** — four `[0-1]` indices from Maria
  Esquivel: land use, water management, pasture management, and an
  overall "all practices" composite.

Joined on farm ID, `r n_distinct(md$Id_gcs)` farms have both, giving
\~`r nrow(read_csv("Derived/Excels/Farm_mgmt_model_data.csv", show_col_types = FALSE) %>% filter(Hill == "shannon"))`
assemblages.

# Why this is hard: the confounding structure

Bird diversity in this region is **mostly a between-ecoregion story** —
the R² of `richness ~ Ecoregion` is $\approx 0.45$. And `Ecoregion` is a
near-perfect stand-in for the environment: elevation and precipitation
are each $\approx 0.82$ explained by which ecoregion a farm sits in
(elevation on its own explains only $\approx 0.15$–$0.22$ of richness —
the diversity signal is regional, not a simple elevational gradient).
The five ecoregions differ in rainfall regime, elevation, forest cover
and biogeographic history, and those differences dominate the diversity
signal.

The four management indices **split in two** with respect to region
(`Scripts/03_exploratory.R`):

- **Land use** and **All practices** are barely region-structured (R² \~
  0.08–0.10) — they carry usable within-region variation.
- **Water management** and **Pasture management** are strongly
  region-structured (R² \~ 0.34–0.35) — they co-vary with whatever else
  varies by region.

So any region-correlated environmental driver of bird diversity that
also correlates with a management index is a confounder. The assumed
causal structure is in `Scripts/dag.R` (@fig-dag): `Ecoregion` is a
stand-in for climate (elevation + precipitation), landscape forest
cover, the regional species pool, and, importantly, also for some farm
management practices (particularly water and pasture management). The
identification strategy is to decompose the ecoregion category into the
underlying *mechanisms* — assess the effect of farm management while
controlling for climate, landscape forest cover, and the regional
species pool.

![Assumed causal DAG behind the analysis (`Scripts/dag.R`). Adjusting for
climate, landscape forest cover and the sampling covariates blocks the
ecoregion backdoors; farmer values and farm size stay
unobserved.](/Figures/DAG.png){#fig-dag width=100%}

The practical obstacle is **collinearity**: elevation, precipitation,
canopy cover, and the farm management practices are all highly
structured by ecoregion.

# The modelling approach

`Scripts/04_farm_mgmt_mod.R` — Bayesian hierarchical regression, one
diversification index at a time:

```         
log(diversity) | resp_se(se, sigma = TRUE) ~ diversification_index_z + <environment> +
    log(Num_pc)_z + Num_hab_z + doy_sin + doy_cos + (1 | farm) + (1 | collector-year)
```

- `resp_se()` propagates each iNEXT estimate's bootstrap /
  coverage-based standard error into the response.
- **Two environmental-adjustment versions** (`Scripts/dag.R`):
  *climate + canopy* — elevation and precipitation each as *x + x²*,
  plus surrounding 10 km canopy cover; *Ecoregion + canopy* — the
  5-level factor as a coarse proxy (robustness check).
- Sampling controls throughout: survey effort `log(Num_pc)`, number of
  habitat types surveyed `Num_hab` (2013 / 2016-17 surveys were
  forest-only, 2019+ spanned the land-use gradient), cyclic day-of-year,
  and farm / collector-year random effects.
- A **no-index baseline** per response, so the variance an index *adds*
  is visible.
- **Primary analysis**: assemblages whose point counts average \< 300 m
  from the farm; the full set is a sensitivity analysis.

Two extensions: a **Piedemonte-only** cut (`Scripts/04c_...`) that
examines diversification just within the Piedemonte region, and a
**canopy scale-of-effect** analysis (`Scripts/06b_...`).

The cyclic day-of-year term deserves a note, since the surveys span the
calendar: `doy_sin` + `doy_cos` together fit one smooth annual sine wave
of any amplitude and phase, with no discontinuity at year-end. Fitted on
the no-index Shannon baseline it is a $\approx 21\%$ peak-to-mean swing,
highest in the Sept–Oct migrant-arrival window (@fig-doy) — real enough
to be worth removing before asking about management.

```{r}
#| label: fig-doy
#| fig-cap: "Fitted seasonal (day-of-year) multiplier on diversity, no-index Shannon baseline. Shaded = Nearctic-migrant season; ticks = the assemblages' mean survey days."
#| fig-width: 9
#| fig-height: 3.4
p_fit + labs(title = NULL)
```

# Preliminary results

## No robust management effect

```{r}
#| label: fig-effects
#| fig-cap: "Standardized effect of a 1-SD increase in each diversification index on log bird diversity (posterior median, 90% CrI), primary analysis. Purple = climate + canopy spec, green = Ecoregion + canopy spec."
#| fig-width: 9
#| fig-height: 4
bare(p_index_effects)
```

Under the **Ecoregion** version every index effect is indistinguishable
from zero (@fig-effects). Under the **climate** version, land use and
all-practices are also $\approx 0$; **pasture and water management are
weakly positive** ($\approx +0.07$ to $+0.08$ for richness and Shannon,
credible interval sometimes excluding zero). The baseline and index R²
are close throughout — the indices add essentially nothing (@fig-r2).

```{r}
#| label: fig-r2
#| fig-cap: "Bayesian R² of the no-index baseline vs the model with each diversification index added (primary analysis, both specs). Adding an index barely moves explained variance."
#| fig-width: 9
#| fig-height: 4
bare(p_bayes_r2)
```

## The species pool is structured by elevation and precipitation

Differences in the diversity estimates could partly reflect differences
in the regional species pool ("how many species could occur here?"). To
check this, `Scripts/01b_species_pool.R` builds a per-farm **potential
species pool** (`spp_pool`): the number of bird species whose Ayerbe
Colombian range polygon contains the farm **and** whose published
elevational range spans the farm's elevation. Elevational limits come
mostly from Suárez-Castro et al. (2024), with some ranges digitised
manually from field guides (Ayerbe-Quiñones 2018, Hilty) — 87 % of the
1 788 range-map species; the rest (Nearctic migrants, seabirds) pass
unfiltered.

But `spp_pool` is itself a composite of the two environmental gradients:
`spp_pool ~ poly(Elev, 2)` gives R² $\approx 0.49$,
`~ poly(precip, 2)` R² $\approx 0.52$, and both together R²
$\approx 0.87$, leaving only $\approx 13\%$ unexplained (@fig-pool-varpart).
Its own association with diversity is an **elevation** effect (strongly
positive alone, $\approx 0$ once `Elev²` is in the model), and adding it
to the management model does not move the focal coefficients
(@fig-env-specs). So a range-map species-pool count carries no
information beyond elevation + precipitation. A composition- or
endemism-weighted pool, or an eBird-based measure, is the next thing to
try if the aim is a *biogeographic* residual.

```{r}
#| label: fig-pool-varpart
#| fig-cap: "Variance of the per-farm range-map species pool partitioned onto the elevation and precipitation gradients; together they explain about 87 percent."
#| fig-width: 9
#| fig-height: 2.6
p_pool_varpart + labs(title = NULL)
```

## The pasture / water signal is precipitation confounding

The two indices that show anything are exactly the two most correlated
with region — and specifically with rainfall. At the farm level,
**`Pasture_mgmt_div` correlates with mean annual precipitation at
r = `r sprintf("%.2f", idx_prec_r["Pasture_mgmt_div"])`** (water
management r = `r sprintf("%.2f", idx_prec_r["Water_mgmt_div"])`; land
use and all-practices are slightly *negative*). Wetter farms have higher
pasture-management diversification scores, and precipitation is a
primary driver of tropical bird diversity — a confounding fork.

`Scripts/04d_species_pool_test.R` refits the model with the
environmental term specified many ways: species pool, elevation² and
precipitation² in every combination that never uses all three (they are
collinear), plus precipitation as a quadratic vs a spline, with and
without elevation co-adjusted. The pasture / water coefficient tracks
**how completely precipitation is adjusted for** (@fig-env-specs) — it
is $\approx +0.08$ to $+0.10$ whenever elevation competes with
precipitation for the shared variance (so precipitation only partly
removes the confound), and falls to $\approx +0.03$ to $+0.05$ with the
credible interval spanning zero whenever precipitation is the sole or
dominant environmental term. Quadratic vs spline makes no difference.

```{r}
#| label: fig-env-specs
#| fig-cap: "Management-index coefficient (pasture, water) across every environmental specification, labelled by the variables it contains and ordered by how completely precipitation is controlled. 10 km canopy and the sampling controls are held fixed throughout. The coefficient shrinks toward zero as precipitation adjustment improves."
#| fig-width: 9
#| fig-height: 6
bare(p_env_specs)
```

## Holding region fixed: Piedemonte only

```{r}
#| label: fig-pied
#| fig-cap: "Piedemonte-only fits (`Scripts/04c`): standardized management-index effect on log diversity, region held fixed, elevation/precipitation/canopy replaced by their first PC. Primary (< 300 m) vs all assemblages. Every credible interval includes zero."
#| fig-width: 9
#| fig-height: 4
bare(p_pied_effects)
```

Restricting to the 25 Piedemonte farms (`Scripts/04c_...`), with the
collinear elevation / precipitation / canopy replaced by their first
principal component, still shows **no convincing management effect**.
This is consistent with the confounding story: management is generally
positive, especially pasture / water management, but every credible
interval includes zero.

## A landscape-context effect: canopy cover scale of effect

```{r}
#| label: tbl-soe
soe %>%
  filter(!is.na(dR2m_canopy), spec %in% c("topography", "topography_farm", "within_eco")) %>%
  slice_max(dR2m_canopy, n = 1, by = c(response, spec)) %>%
  transmute(
    Response = recode(str_remove(response, "_q[0-9]"), !!!hill_lab),
    Spec = recode(spec, topography = "+ elevation + precip", topography_farm = "+ topo + farm RE",
                  within_eco = "within-ecoregion (Mundlak)"),
    `Radius (m)` = radius_m,
    `Std. effect [95% CI]` = sprintf("%+.3f [%+.3f, %+.3f]", canopy_beta, canopy_lo, canopy_hi)
  ) %>%
  arrange(Response, Spec) %>%
  kable(caption = "Canopy-cover scale of effect (Scripts/06b): net of elevation + precipitation, bird richness is positively associated with surrounding woody-vegetation cover, strongest at a broad ~8-9 km radius. This is a landscape-context effect, not a farm-management one.")
```

## Specification robustness

Three modelling choices were checked (`Scripts/04e_spec_checks.R`); none
changes the conclusion.

**`(1 | CollectorXyear)`** — the \[dataset × year-group\] random effect
(8 levels) is only partly confounded with `Ecoregion` (Cramér's V =
0.38; the three largest datasets each span 4–5 ecoregions). Its standard
deviation is 0.18–0.31 on the log scale, so it captures real
field-method differences among the six source datasets. Dropping it
shifts every management coefficient by $\le 0.02$, on the full data and
in Piedemonte alike — it is not doing region-confounding control.

**`resp_se()`** — propagating the iNEXT standard errors makes the fit a
precision-weighted regression. Dropping it (treating the point estimates
as exact) gives \~25 % *larger* management coefficients and \~7 % wider
credible intervals, because the well-sampled assemblages — which show a
weaker signal — then count for less. `resp_se` is both the honest and
the conservative choice.

**Distance-to-farm cutoff** — for some assemblages the point counts
average well off the nominal farm (@fig-dist): different sampling teams
placed counts along access roads or on neighbouring land. The primary
analysis keeps only assemblages whose counts average \< 300 m from the
farm; using all assemblages instead shifts the coefficients by
$\le 0.03$, toward zero if anything.

```{r}
#| label: fig-dist
#| fig-cap: "Mean distance of an assemblage's point counts from its farm. Dashed line = the 300 m primary-analysis cutoff; the long right tail is dropped from the primary set."
#| fig-width: 9
#| fig-height: 2.8
read_csv("Derived/Excels/Farm_mgmt_model_data.csv", show_col_types = FALSE) %>%
  filter(Hill == "shannon", !is.na(dist_farm)) %>%
  distinct(Assemblage, dist_farm) %>%
  ggplot(aes(dist_farm)) +
  geom_histogram(binwidth = 100, boundary = 0, fill = "grey70", colour = "white") +
  geom_vline(xintercept = 300, linetype = "dashed", colour = "#d95f02") +
  labs(x = "Mean distance of the point counts from the farm (m)", y = "Assemblages") +
  theme_minimal(11)
```

```{r}
#| label: tbl-scheck
scheck %>%
  filter(env == "climate") %>%
  mutate(variant = recode(variant, base = "base", no_collector = "no CollectorXyear RE",
                          no_resp_se = "no resp_se"),
         hill = recode(hill, !!!hill_lab), index = recode(index, !!!index_lab)) %>%
  select(hill, index, variant, focal_est) %>%
  pivot_wider(names_from = variant, values_from = focal_est) %>%
  arrange(hill, index) %>%
  kable(digits = 3,
        caption = "Focal coefficient (posterior median) under the climate spec: base vs dropping the CollectorXyear random effect vs dropping resp_se. Primary analysis.")
```

# Interpretation

**At this scale, with these data, farm management diversification shows
no detectable effect on bird taxonomic diversity.**

- The **land-use** and **all-practices** indices — the ones separable
  from region — are flat null under every specification.
- The **pasture** and **water-management** indices show a weak positive
  coefficient under the parametric climate adjustment, but it is a
  **precipitation-confounding artifact**: it appears only when
  precipitation is weakly controlled and disappears (credible interval
  spanning zero) whenever precipitation is adjusted well — including
  under the `Ecoregion` factor, which is R² 0.83 with precipitation.
- The environmental gradients dominate, and the region-bound indices are
  too collinear with them for a management effect to be identified from
  observational data.
- Separately, **surrounding landscape forest cover** (\~8–9 km) *is*
  positively associated with bird richness, net of climate — a
  landscape-context effect that is not about on-farm management.

## Why `Ecoregion` is legitimate adjustment, not over-adjustment

The concern with the 5-level factor was that it might absorb
region–management covariation that is not confounding — e.g. if the SCR
programme promoted water-management practices more in wetter regions as
policy, conditioning on `Ecoregion` would regress away part of a real
effect. Three things argue against that:

1.  **The attenuation is reproducible with no region factor at all**,
    purely by making precipitation the dominant environmental term
    (dropping the elevation term from the continuous spec). Continuous
    climate variables do not carry a "the factor soaks up management
    structure" mechanism, so if the `Ecoregion` result were
    over-adjustment they could not reproduce it — but they do.
2.  **There is a concrete confounding path**:
    `Pasture_mgmt_div ~ precipitation` r = 0.30 (and only 0.02 with
    elevation), and precipitation drives bird diversity. The DAG has
    this fork.
3.  **The `climate` spec controls precipitation *worse* than the factor,
    not better.** Elevation and precipitation go in as two
    negatively-correlated continuous terms; because they trade off,
    precipitation removes less of the pasture / water confound than it
    would alone. Drop the elevation term — or use the single flexible
    `Ecoregion` factor — and precipitation absorbs the full confound:
    the coefficient falls to \~0.04 with the credible interval spanning
    zero, regardless of whether precipitation enters as a quadratic or a
    spline. So the `climate`–`ecoregion` gap is the parametric climate
    spec *under-adjusting* precipitation under collinearity, not the
    factor over-adjusting.

**Caveat.** With everything this collinear, "legitimate adjustment" is
an inference from the pattern, not a proof. Precipitation and "regional
programme rollout" may themselves be confounded, so we cannot fully
separate "wetter, so more birds" from "wetter region, so more
water-management promotion". But under either reading the direction is
the same: the well-precipitation-adjusted near-zero estimate is the one
to report, and the +0.08 is not a defensible management effect.

# Where this goes

- **A better environmental term in the `climate` spec** — the parametric
  `Elev + Elev² + precip + precip²` under-controls precipitation because
  the two collinear axes trade off (§5.1). A single flexible term (an
  `s(lat, long)` spatial spline, or dropping the elevation quadratic),
  or simply taking the `Ecoregion`-factor result as primary, would close
  the `climate`–`ecoregion` gap at \~0.
- **Other response axes** — functional or phylogenetic diversity, or
  species-of-conservation-concern subsets, may respond to management
  where taxonomic Hill numbers do not.
- **The biogeographic residual** — an endemism-weighted pool or an eBird
  composition measure, if the compositional part of the `Ecoregion`
  effect is worth chasing.
- **Farm isolation** — `dist_predio_cercano` (distance to the nearest
  neighbouring property, from the MJE spreadsheet) is uncorrelated with
  the point-count `Distancia_farm` ($r \approx 0$), so it is a genuinely
  new covariate: a candidate predictor and/or a spatial-autocorrelation
  control.
- **Manuscript framing** — the honest story is "farm-scale management
  diversification does not detectably raise bird taxonomic diversity
  once the environmental gradient is accounted for; surrounding
  landscape forest cover does" — reported with the collinearity and
  observational caveats.
- Better understanding of iNEXT diversification indices - Consideration
  of alternative cutoff values.

# Appendix: methodological questions

Six questions that came up while building the analysis, with the answers
that shaped the current specification.

## Is canopy cover "required by the DAG"?

Not as a farmer-values backdoor — that was a scale mismatch. The canopy
covariate is woody cover in a $\approx 10$ km buffer, set by topography,
protected areas and hundreds of landholders; no single farmer's
priorities move it. So there is no farmer-values-to-canopy arrow and
canopy is not blocking a farmer-values path. What canopy legitimately
does: (1) in the `climate` spec it blocks the regional backdoor
FarmDiv \<- Ecoregion -\> LandForest -\> BirdDiv; (2) everywhere, it is a
strong predictor of diversity, so it reduces residual variance. Under
the `Ecoregion` spec it is not needed at all but is harmless and
variance-reducing, so it stays in both. The genuine residual concern —
bird-minded farmers doing *on-farm* things the four indices miss (snag
retention, less pesticide, wider fencerows) — is a farm-scale path that
canopy cannot proxy; it is left as an explicit unmeasured-confounder
caveat in `Scripts/dag.R`.

## Why is farm size in the DAG?

Farm area (unmeasured) has three arrows. FarmSize -\> FarmDiv: a bigger
farm has more room and more management units, so more opportunity to
score high on indices that are partly counts of practices.
FarmSize -\> NumPC: bigger farms get more point counts.
FarmSize -\> BirdDiv: species–area — more area means larger populations,
more interior habitat, proportionally less edge. Conditioning on
`NumPC` (already in every model) closes the
FarmDiv \<- FarmSize -\> NumPC -\> BirdDiv path — this is why `NumPC` is
now a genuine confounder in the DAG's minimal adjustment set, not just a
precision term. The direct FarmSize -\> BirdDiv path is **not** closed
by anything in the model and would bias a true null upward. Measuring
farm hectares would close it; it is on the re-analysis checklist.

## `resp_se(se, sigma = TRUE)` vs `sigma = FALSE`

With `sigma = TRUE` the total variance for assemblage *i* is
$se_i^2 + \sigma^2$ — the known iNEXT bootstrap SE **plus** an estimated
residual SD shared across rows. `sigma = FALSE` uses only $se_i^2$
(pure inverse-variance / meta-analysis weighting), which assumes the
known SEs are the whole error story. Refitting pasture × richness
(climate primary) both ways: `TRUE` gives
$+0.081\ [-0.002,\ 0.164]$, `FALSE` gives
$+0.088\ [-0.007,\ 0.184]$ (a slightly larger point estimate and a
$\approx 15\%$ wider interval). The `TRUE` model's estimated residual SD
is $\approx 0.15$, larger than the mean known measurement error
($\overline{se_{\log}} \approx 0.09$) — there is real
assemblage-to-assemblage scatter beyond the bootstrap SE, so
`sigma = TRUE` is the correct choice. (This is separate from the
"drop `resp_se` entirely" check in §4.6, which also discards the known
SEs and gives $\approx 25\%$ larger coefficients.)

## Removing the $\pm 250$ m species-pool elevation buffer

Done. A farm's elevation must now fall inside a species' published
elevational range exactly. This keeps 96 % of the range-polygon hits
(was 98 %); `spp_pool ~ climate` R² and the management coefficients in
@fig-env-specs are unchanged.

## What does it mean if Bayesian R² *decreases* when an index is added?

Almost entirely a subset artifact, not a sign the index hurts. Bayesian
R² is computed on the rows each model was fit on. The no-index baseline
uses all assemblages with complete covariates ($n \approx 89$–$98$); the
water- and pasture-management models drop farms missing those indices
($n \approx 73$–$80$). Different data, so the R² values are not directly
comparable — removing $\approx 20$ rows changes both the fitted-value
variance and the residual variance. Two smaller contributors: the
`normal(0, 0.75)` prior shrinks a near-zero coefficient toward 0, so an
uninformative index adds almost nothing to explained variance; and for
Shannon/Simpson the water/pasture *subset* actually has a **higher** R²
than the baseline, only richness is slightly lower — that inconsistency
across responses is the signature of subset noise. A clean comparison
would refit the baseline on the matched $n$ (backlog).

## Would LOO for the nuisance terms change anything now?

Not at this stage. With $n \approx 73$–$98$ and the regularising
`normal(0, 0.75)` prior, `Scripts/04e_spec_checks.R` already shows that
dropping a nuisance block (the CollectorXyear RE, and by extension the
day-of-year terms) moves the focal coefficient by $\le 0.02$. LOO would
formalise "keep them" without changing the conclusion, and LOO with this
many single-assemblage farms and the measurement-error structure needs
care (Pareto-$k$ diagnostics, possibly moment-matching). It is worth
doing when the specifications are being finalised for the manuscript on
a larger dataset, not now.
