---
title: "Piedemonte-only analysis: bird diversity and farm management diversification"
subtitle: "`Scripts/04c_farm_mgmt_piedemonte.R`, `05c_farm_mgmt_piedemonte_plot.R`"
date: today
format:
  pdf:
    toc: false
    number-sections: true
    fig-pos: "H"
    geometry: margin=1in
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

source("Scripts/05c_farm_mgmt_piedemonte_plot.R")   # -> p_pied_effects

library(tidyverse)
library(knitr)

hill_lab  <- c(richness = "Richness (q = 0)", shannon = "Shannon (q = 1)", simpson = "Simpson (q = 2)")
index_lab <- c(baseline = "(baseline)", Land_use_div = "Land use", Water_mgmt_div = "Water mgmt",
               Pasture_mgmt_div = "Pasture mgmt", All_practices_div = "All practices")

mm <- read_csv("Derived/Excels/Farm_mgmt_piedemonte_summaries.csv", show_col_types = FALSE)
pd <- read_csv("Derived/Excels/Farm_mgmt_piedemonte_data.csv", show_col_types = FALSE)

pf <- pd %>% distinct(Id_gcs, .keep_all = TRUE)
n_farm <- nrow(pf)
n_asm  <- pd %>% filter(Hill == "shannon") %>% nrow()
n_300  <- pd %>% filter(Hill == "shannon", dist_farm < 300) %>% nrow()
env_cor <- cor(pf %>% select(Elev_mean, Tot_prec_mean, canopy_10k))
env_pca <- prcomp(pf %>% select(Elev_mean, Tot_prec_mean, canopy_10k), scale. = TRUE)
pc1_var <- summary(env_pca)$importance["Proportion of Variance", "PC1"]
```

# The question

Across all five ecoregions (`Scripts/04_farm_mgmt_mod.R`), no farm-management
diversification index shows a robust association with bird diversity. The only
non-null coefficients — pasture and water management, weakly positive under the
climate-adjusted model — vanish when `Ecoregion` is added, and those two indices
are the ones most confounded with region. The causal DAG (`Scripts/dag.R`)
identifies the **regional species pool** as the one confounder the climate model
cannot fully capture with elevation, precipitation and canopy.

Restricting the analysis to a **single ecoregion** removes that by design: the
regional species pool is held approximately constant, the `Ecoregion → index`
pathway is gone, and between-region climate structure no longer contributes. If a
real management effect exists, this is where it should be easiest to see; a null
here is the most credible null in the analysis.

**Piedemonte** is the natural choice — `r n_farm` of the 69 matched farms,
roughly half the point counts in the source dataset, and the full range of the
management indices.

# Data

```{r}
#| label: tbl-data

pd %>%
  filter(Hill == "shannon") %>%
  summarize(
    Farms = n_distinct(Id_gcs),
    Assemblages = n(),
    `Assemblages < 300 m` = sum(dist_farm < 300),
    `Single-assemblage farms` = sum(table(Id_gcs) == 1)
  ) %>%
  pivot_longer(everything(), names_to = "Quantity", values_to = "n") %>%
  kable(caption = "Piedemonte subset.")
```

```{r}
#| label: tbl-index-cov

pf %>%
  summarize(across(c(Land_use_div, All_practices_div), ~ sum(!is.na(.x)))) %>%
  pivot_longer(everything(), names_to = "index", values_to = "farms") %>%
  bind_rows(tibble(index = c("Water_mgmt_div", "Pasture_mgmt_div"),
                   farms = mm %>% filter(term == "focal_z", index %in% c("Water_mgmt_div", "Pasture_mgmt_div"),
                                         data_subset == "primary", re_spec == "full_re") %>%
                     distinct(index, n_farm) %>% pull(n_farm))) %>%
  mutate(index = recode(index, !!!index_lab)) %>%
  arrange(desc(farms)) %>%
  kable(col.names = c("Index", "Farms with data"),
        caption = "Management-index coverage (of 25 farms). Water and pasture management take the missing-data hit -- their models are n = 17 and should be read with wide credible intervals.")
```

# Model

## The environmental gradient

Within Piedemonte, elevation, precipitation and 10 km canopy cover are almost
perfectly collinear — they form a single foothill gradient (lower = drier, less
forested; higher = wetter, more forested):

```{r}
#| label: tbl-collin

m <- round(env_cor, 2)
m[lower.tri(m, diag = TRUE)] <- NA
as_tibble(m, rownames = "var") %>%
  kable(caption = "Farm-level correlations among the three environmental covariates, within Piedemonte (n = 25).")
```

They cannot go in a 25-farm model separately, so they are replaced by
**`env_pc1`**, the first principal component of the three, which captures
**`r sprintf("%.0f%%", 100 * pc1_var)`** of their joint variance (loadings
`r paste(sprintf("%.2f", env_pca$rotation[, "PC1"]), collapse = " / ")` for
elevation / precipitation / canopy — the three load equally). `env_pc1` is linear
only; over an elevation range of ~270–760 m the relationship is monotonic and
25 farms will not support a quadratic.

## Specification

Same Bayesian measurement-error structure as the full analysis, one index at a
time:

```
log(diversity) | resp_se(se, sigma = TRUE) ~
    index_z + env_pc1_z +
    Num_pc_log_z + Num_hab_z + doy_sin + doy_cos +
    (1 | Id_gcs) + (1 | CollectorXyear)
```

- Response uncertainty (the iNEXT bootstrap / coverage-based SE) enters through
  `resp_se()`. All three Hill numbers.
- `Num_hab` (number of habitat types surveyed) is a sampling-scope control: 2013
  and 2016/17 surveys covered forest only, 2019 onward the land-use gradient.
- A **no-index baseline** per response, for the variance comparison.
- **No "ecoregion" model version** — region is fixed, nothing to bracket against.
- All continuous predictors re-standardised *within* Piedemonte, so a coefficient
  is the effect of a 1-SD change among Piedemonte farms.
- Primary analysis = assemblages whose point counts average < 300 m from the farm
  (`r n_300` of `r n_asm`); the full set is a sensitivity.
- Priors as in `04`; `adapt_delta = 0.999`, `iter = 4000` (small n).

## Sensitivities

1. **Full assemblage set** (no 300 m cutoff).
2. **Drop `(1 | CollectorXyear)`** — all 8 collector-year batches are present
   within Piedemonte, so this is the natural place to check whether that grouping
   re-absorbs a management signal.

# Results

All 39 fits converged (max R-hat `r sprintf("%.3f", max(mm$max_rhat))`,
`r sum(mm$n_divergent)` divergences total).

## The management-diversification coefficient

```{r}
#| label: fig-effects
#| fig-width: 9
#| fig-height: 4.2

p_pied_effects + labs(title = NULL, subtitle = NULL)
```

```{r}
#| label: tbl-focal

mm %>%
  filter(term == "focal_z", data_subset == "primary", re_spec == "full_re") %>%
  mutate(cell = sprintf("%+.3f [%+.3f, %+.3f]  P=%.2f  (n=%d)", estimate, conf_low, conf_high, p_direction_pos, n_farm),
         hill = recode(hill, !!!hill_lab), index = recode(index, !!!index_lab)) %>%
  select(hill, index, cell) %>%
  pivot_wider(names_from = hill, values_from = cell) %>%
  kable(caption = "Standardized effect of a 1-SD increase in each index on log diversity (primary analysis): posterior median [90% CrI], P(effect > 0), and n farms.")
```

Every 90% credible interval includes zero. **All practices and Land use** — the
indices least confounded with the environmental gradient — are flat null across
all three Hill numbers. **Pasture** (+0.05 to +0.08) and **water** (+0.05 to
+0.06) management are weakly positive with the same sign as the full analysis,
but with n = 17 farms the intervals are wide; water management on richness is the
closest to a signal (P = 0.94).

## Variance explained

```{r}
#| label: tbl-r2

mm %>%
  filter(data_subset == "primary", re_spec == "full_re") %>%
  distinct(hill, index, bayes_R2) %>%
  mutate(hill = recode(hill, !!!hill_lab), index = recode(index, !!!index_lab)) %>%
  pivot_wider(names_from = index, values_from = bayes_R2) %>%
  kable(digits = 2, caption = "Bayesian R^2: no-index baseline vs each index model. The indices add little; the small gains for water / pasture are on n = 17.")
```

## Sensitivity

```{r}
#| label: tbl-sens

mm %>%
  filter(term == "focal_z") %>%
  mutate(fit = case_when(data_subset == "full" ~ "Full set",
                         re_spec == "no_collector" ~ "No collector RE",
                         TRUE ~ "Primary"),
         val = sprintf("%+.3f", estimate),
         hill = recode(hill, !!!hill_lab), index = recode(index, !!!index_lab)) %>%
  select(hill, index, fit, val) %>%
  pivot_wider(names_from = fit, values_from = val) %>%
  arrange(hill, index) %>%
  kable(caption = "Focal coefficient (posterior median) across the three fits. The estimates barely move -- dropping the collector random effect changes nothing, so it was not masking a management effect.")
```

# Interpretation

- **The null holds in the strongest test available.** With region fixed, the
  species pool held approximately constant, and the full range of the
  region-separable indices (Land use, All practices) present, there is still no
  detectable management–diversity association.
- The **weak pasture / water pattern** survives into Piedemonte at the same
  magnitude, but it never separates from zero here (n = 17) any more than it did
  in the full analysis. It is consistent with a small real effect *or* with
  residual within-Piedemonte confounding; this analysis cannot distinguish them.
- **The collector random effect is exonerated** — the concern that
  `(1 | CollectorXyear)` was soaking up a management signal (because 4 of 6
  collectors are Piedemonte-only) does not hold: removing it leaves every
  coefficient essentially unchanged.

# Caveats

- **Power.** Effectively a ~`r n_farm`-farm regression (most farms contribute one
  assemblage) with a measurement-error response. It can rule out large effects
  (standardized |β| greater than roughly 0.15–0.2) but not small ones. A null
  means "no evidence of an effect at this power", not "evidence of no effect".
- **Single environmental axis.** `env_pc1` is one summary of a three-way
  collinear gradient; the individual roles of elevation, rainfall and canopy
  cannot be separated within Piedemonte.
- **Observational.** Unmeasured farmer priorities / farm size and within-
  Piedemonte environment beyond `env_pc1` remain possible confounders — this is a
  cleaner comparison than the full analysis, not a randomized one.
