---
title: "Proposal: a Piedemonte-only cut of the bird-diversity ~ farm-management analysis"
subtitle: "Companion to `Scripts/04_farm_mgmt_mod.R` and `Scripts/dag.R`"
date: today
format:
  pdf:
    toc: false
    number-sections: true
    fig-pos: "H"
    geometry: margin=1in
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup

library(tidyverse)
library(knitr)
library(cowplot)
ggplot2::theme_set(theme_cowplot(11))

## Same modelling frame 04 builds, plus the per-assemblage distance-to-farm 04 uses for the cutoff.
d <- read_csv("Derived/Excels/Farm_mgmt_model_data.csv", show_col_types = FALSE)

we <- "../Ssp-bird-data-wrangling/Derived/Excels/"
asm_dist <- read_csv(paste0(we, "Site_covs.csv"), show_col_types = FALSE) %>%
  select(Id_muestreo_no_dc, Id_gcs, Distancia_farm) %>%
  left_join(read_csv(paste0(we, "Event_covs.csv"), show_col_types = FALSE), by = "Id_muestreo_no_dc") %>%
  filter(!is.na(Ano)) %>%
  mutate(Assemblage = str_replace_all(paste(Uniq_db, Id_gcs, Ano_grp, Season, sep = "."), " |-", "_")) %>%
  summarize(dist_farm = mean(Distancia_farm, na.rm = TRUE), .by = Assemblage)

pied <- d %>% filter(Ecoregion == "Piedemonte") %>% left_join(asm_dist, by = "Assemblage")
pied_farm <- pied %>% distinct(Id_gcs, .keep_all = TRUE)

indices <- c("Land_use_div", "Water_mgmt_div", "Pasture_mgmt_div", "All_practices_div")
index_lab <- c(Land_use_div = "Land use", Water_mgmt_div = "Water mgmt",
               Pasture_mgmt_div = "Pasture mgmt", All_practices_div = "All practices")

n_farm <- nrow(pied_farm)
n_asm_q12 <- pied %>% filter(Hill == "shannon") %>% nrow()
n_asm_q0  <- pied %>% filter(Hill == "richness") %>% nrow()
n_asm_300 <- pied %>% filter(Hill == "shannon", dist_farm < 300) %>% nrow()

## within-Piedemonte environmental collinearity + PC1
env <- pied_farm %>% select(Elev_mean, Tot_prec_mean, canopy_10k) %>% drop_na()
env_cor <- cor(env)
env_pca <- prcomp(env, scale. = TRUE)
pc1_var <- summary(env_pca)$importance["Proportion of Variance", "PC1"]
pied_farm <- pied_farm %>%
  mutate(env_pc1 = as.numeric(predict(env_pca, newdata = select(pied_farm, Elev_mean, Tot_prec_mean, canopy_10k))[, 1]))

idx_pc1_cor <- sapply(indices, function(v) cor(pied_farm[[v]], pied_farm$env_pc1, use = "pairwise"))
```

# Why do it

Every version of the full-data analysis is dominated by *between-ecoregion*
differences. Bird diversity is mostly a regional story (R² of
`richness ~ Ecoregion` ≈ 0.45), the two management indices with a signal
(pasture, water) are the ones most confounded with region, and that signal
disappears the moment `Ecoregion` is adjusted for. The causal DAG
(`Scripts/dag.R`) says the one thing the primary "climate" version cannot
capture is the **regional species pool** — biogeographic history that elevation,
precipitation and canopy do not encode.

Restricting to a single ecoregion removes all of that by design:

- the regional species pool is held (approximately) constant — the eBird
  species-pool proxy on the backlog becomes unnecessary for this cut;
- the `Ecoregion -> FarmDiv` path and all between-region climate structure are
  gone;
- what remains to worry about: unobserved farmer values / farm size, the
  *within*-Piedemonte environment, and the data-collector / protocol confound.

**Piedemonte** is the natural choice: it holds `r n_farm` of the 69 matched
farms, roughly half of all point counts in the source data, and spans the full
range of the management indices. If the region-separable indices show nothing
here, that is the most credible null in the whole analysis; if they show
something, it is the strongest signal.

# Feasibility

```{r}
#| label: tbl-feas

tibble(
  Quantity = c("Matched farms", "Assemblages (q = 1 / q = 2)",
               "Assemblages (q = 0 richness)", "Assemblages surviving the < 300 m cutoff",
               "CollectorXyear levels present", "Farms with a single assemblage"),
  Value = c(as.character(n_farm),
            as.character(n_asm_q12),
            as.character(n_asm_q0),
            paste0(n_asm_300, " of ", n_asm_q12),
            as.character(n_distinct(pied$CollectorXyear)),
            as.character(sum(count(filter(pied, Hill == "shannon"), Id_gcs)$n == 1)))
) %>%
  kable(caption = "Piedemonte subset size.")
```

```{r}
#| label: tbl-index-cov

pied_farm %>%
  summarize(across(all_of(indices), ~ sum(!is.na(.x)))) %>%
  pivot_longer(everything(), names_to = "index", values_to = "farms_with_data") %>%
  mutate(index = recode(index, !!!index_lab),
         range = sapply(indices, function(v) {
           x <- pied_farm[[v]]; sprintf("%.1f - %.1f", min(x, na.rm = TRUE), max(x, na.rm = TRUE))
         })) %>%
  kable(col.names = c("Index", "Farms with data", "Range"),
        caption = "Management-index coverage within Piedemonte (of 25 farms). Water and pasture management take the missing-data hit.")
```

Land use and All practices are usable across nearly all farms and span their full
`[0-1]` range. Water and pasture management are only available for
`r sum(!is.na(pied_farm$Water_mgmt_div))` farms — analyses of those two will have
wide credible intervals and should be read cautiously.

# The wrinkle: within Piedemonte the environment collapses to one axis

In the full dataset, canopy cover is only moderately correlated with elevation
(r ≈ 0.44), so elevation, precipitation and canopy can all sit in the model
separately. **Within Piedemonte they do not:**

```{r}
#| label: tbl-collin

env_cor[lower.tri(env_cor, diag = TRUE)] <- NA
as_tibble(round(env_cor, 2), rownames = "var") %>%
  kable(caption = "Farm-level correlations among the three environmental covariates, within Piedemonte (n = 25).")
```

Piedemonte is a foothill gradient — lower farms are drier and less forested,
higher farms wetter and more forested, all one thing. Putting all three in a
25-farm model would make them individually unidentifiable.

**Proposed fix:** replace them with the first principal component of
{elevation, precipitation, canopy}, which captures
**`r sprintf("%.0f%%", 100 * pc1_var)`** of their joint variance — a single
"environmental gradient" control.

```{r}
#| label: fig-pc1
#| fig-width: 9
#| fig-height: 3.4
#| fig-cap: "Left: the three covariates collapse onto one gradient (PC1). Right: All practices is orthogonal to that gradient (the clean test); Land use is mildly correlated."

long_env <- pied_farm %>%
  select(env_pc1, Elev_mean, Tot_prec_mean, canopy_10k) %>%
  pivot_longer(-env_pc1) %>%
  mutate(name = recode(name, Elev_mean = "Elevation (m)", Tot_prec_mean = "Precip (mm)", canopy_10k = "Canopy 10 km (%)"))
p_a <- ggplot(long_env, aes(env_pc1, value)) +
  geom_point(size = 1.3, alpha = 0.8) +
  facet_wrap(~name, scales = "free_y") +
  labs(x = "Environmental PC1", y = NULL)

long_idx <- pied_farm %>%
  select(env_pc1, Land_use_div, All_practices_div) %>%
  pivot_longer(-env_pc1) %>%
  mutate(name = recode(name, Land_use_div = sprintf("Land use (r = %.2f)", idx_pc1_cor["Land_use_div"]),
                       All_practices_div = sprintf("All practices (r = %.2f)", idx_pc1_cor["All_practices_div"])))
p_b <- ggplot(long_idx, aes(env_pc1, value)) +
  geom_point(size = 1.3, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, colour = "#d95f02", linewidth = 0.6) +
  facet_wrap(~name) +
  labs(x = "Environmental PC1", y = "Index [0-1]")

plot_grid(p_a, p_b, nrow = 1, rel_widths = c(1.5, 1))
```

# Proposed model

Same Bayesian measurement-error structure as `04`, one index at a time:

```
log(diversity) | resp_se(se, sigma = TRUE) ~
    index_z + env_pc1_z +
    Num_pc_log_z + Num_hab_z + doy_sin + doy_cos +
    (1 | Id_gcs) + (1 | CollectorXyear)
```

- Priors, `adapt_delta = 0.99`, and the coverage-based q = 0 SE all as in `04`.
- A **no-index baseline** per response, for the `bayes_R2` comparison.
- **No "ecoregion" robustness version** — region is fixed, nothing to bracket
  against.
- `env_pc1` linear only: over an elevation range of ~270-760 m the relationship
  is monotonic, and 25 farms will not support a quadratic.
- The `< 300 m` distance-to-farm cutoff carries over
  (`r n_asm_300` of `r n_asm_q12` assemblages), with the full set as a
  sensitivity as in `04`.

## Indices and responses

| Index | Farms | Note |
|---|---|---|
| **All practices** | `r sum(!is.na(pied_farm$All_practices_div))` | **primary** — orthogonal to the environmental gradient (r = `r sprintf("%.2f", idx_pc1_cor["All_practices_div"])`) |
| Land use | `r sum(!is.na(pied_farm$Land_use_div))` | secondary — mildly correlated with the gradient (r = `r sprintf("%.2f", idx_pc1_cor["Land_use_div"])`) |
| Water / pasture mgmt | `r sum(!is.na(pied_farm$Water_mgmt_div))` | thin — report with a wide-CrI caveat |

All three Hill numbers (q = 0 richness has `r n_asm_q0` assemblages).

## Sensitivity checks

- **Drop `(1 | CollectorXyear)`.** This is the natural place to run the backlog
  question of whether the collector random effect re-absorbs a management signal:
  all 8 collector-year levels are present within Piedemonte, and the
  forest-only (2013 / 2016-17) vs land-use-gradient (2019+) sampling change is
  the main thing that grouping carries.
- **Full distance set** vs the `< 300 m` primary, as in `04`.
- Optionally **`env_pc1` replaced by elevation alone** (or canopy alone) — given
  All practices is orthogonal to the gradient, that result should be robust to
  the choice.

# What this can and cannot tell us

Effectively a ~`r n_farm`-point regression (most farms contribute one
assemblage) with a measurement-error response. It can **rule out large effects**
(standardized |beta| greater than roughly 0.15-0.2) but not small ones. A null
result means "no evidence of an effect at this power", not "evidence of no
effect". Unobserved farmer values / farm size and the within-Piedemonte
environment beyond PC1 remain possible confounders — this is a cleaner
observational comparison, not a randomized one.

# Implementation

Two options:

1. **A `region_subset` dimension in `04`'s grid** (`"all"` / `"piedemonte"`),
   swapping the region block for `env_pc1_z` and filtering
   `Ecoregion == "Piedemonte"` when `piedemonte`. ~12 extra fits, a few minutes.
   Keeps everything in one script but adds branching.
2. **A standalone `Scripts/04c_farm_mgmt_piedemonte.R`** that sources `04`'s data
   build and refits the reduced grid. Cleaner separation given `04` is already
   dense; a small `05`-style plot script and a short report section would follow.

Recommend option 2.
